---
title: "code_Ecocapsules"
output: html_document
date: "2026-01-11"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
#load packages
install.packages("pacman")

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ComplexHeatmap", force=TRUE)


library(pacman)
pacman::p_load(data.table,tidyverse,stringr, ggpubr, tools,patchwork, ComplexHeatmap,openxlsx, readxl,janitor, viridis,magrittr,dendsort,vegan,lubridate, RColorBrewer,scales,ggrepel,ggbeeswarm, ggExtra, patchwork,DIZtools, tidytext, circlize, writexl, lemon, vegan)
```

Here I'm also using the updated metadata (manual curation from google doc https://docs.google.com/spreadsheets/d/1bE1L_YkGLn-OSozm9H0W8fleZvGNJIpHnOwosJVOE60/edit?gid=284466228#gid=284466228)

#dirs
```{r}
home_dir = "~/Documents/ETH/Ecoli_capsules/Ecoli_capsules_analysis/Ecoli_capsules_diversity/"

inputdir = paste0(home_dir,"inputdir/")
outputdir = paste0(home_dir,"outputdir/")
plotdir = paste0(home_dir,"plotdir/")

```

#--------PREPARE DATA---------

## NCBI collection

Load ktyps output, assign K positive, negative or partial depending on completeness of KpsEDCSMT and K-type-specific genes and remove unused columns for this analysis.
```{r}
ncbi_short = fread(paste0(inputdir,"ncbi_flank_20260108_ktyps.tsv"), header = T)  %>%
  dplyr::distinct(genome_id, predicted, pred_nr_genes,
                  pred_genes_in_genome,pred_acc_bitscore,pred_is_complete, KpsEDCSMT_nr_genes,
                  KpsEDCSMT_genes_in_genome,KpsEDCSMT_acc_bitscore, KpsEDCSMT_is_complete,
                  KpsFU_nr_genes,KpsFU_genes_in_genome, KpsFU_acc_bitscore,KpsFU_is_complete) %>%
  mutate(manual_ktyp = case_when(pred_is_complete == 1 & KpsEDCSMT_is_complete == 1 ~ "K_pos",
                                 pred_is_complete ==0 & KpsEDCSMT_is_complete == 0 ~ "K_neg",
                                 TRUE ~ "K_partial")) 

# check these are correctly assigned
ncbi_short %>% distinct(manual_ktyp, pred_is_complete,KpsEDCSMT_is_complete)
# quick check of KpsC unidentified  
table(ncbi_short$manual_ktyp, ncbi_short$predicted)
# all no KpsC identified are K-neg -

# No K-type assigned is always negative
  
# make this lighter for all downstream analyses
ncbi_short=ncbi_short %>%
  dplyr::select(genome_id, predicted, manual_ktyp)

```


### Annotate with K-type groups
```{r}
Kgroups <- readxl::read_xlsx(paste0(inputdir,"subgroups.xlsx"))

ncbi_short = ncbi_short %>%
  left_join(Kgroups, by=c("predicted"="new_name")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           TRUE ~ group))

```

### Annotate with 99% ANI representatives
```{r}
skani = fread(paste0(inputdir,"ani98_ani99_representatives.tsv"), sep = "\t", header=T) %>%
    mutate(genome_id = gsub("\\.\\d+$", "", genome))

ncbi_short = ncbi_short %>% left_join(skani)
```

### Annotate with metadata, ST, phylogroup, O- and H-type

```{r}
metad = readxl::read_xlsx(paste0(inputdir,"ncbi_collection_curatedmetadata_260112.xlsx"))

# deprecated and duplicated genomes were removed in this file so this is the set of 32043 genomes to keep

ncbi_short = ncbi_short %>% inner_join(metad, by="genome_id")

```

### Annotate with ST, phylogroup and O and H type from ectyper

```{r}
#load phylogroups, ectyper and mlst
phylogroups = fread(paste0(inputdir,"assigned_phylogroups.csv"))
phylogroups[duplicated(phylogroups$genome_id), ]
n_distinct(phylogroups$genome)
# 39381 genomes

phylogroups = phylogroups %>%
  mutate(genome_id = sapply(strsplit(as.character(genome),"\\."),'[',1)) %>%
  dplyr::select(-genome)


ectyper = fread(paste0(inputdir,"ectyper_ecoli_metadata.tsv")) %>%
  filter(Name != "Name")
ectyper[duplicated(ectyper$Name), ]
n_distinct(ectyper$Name)
# 39460 genomes
ectyper = ectyper %>%
  mutate(genome_id = sapply(strsplit(as.character(Name),"\\."),'[',1)) %>%
  dplyr::select(-Name)


mlst = fread(paste0(inputdir,"mlst_results_combined.tsv"))
mlst[duplicated(mlst$V1), ]
n_distinct(mlst$V1)
# 39460 genomes
mlst = mlst %>%
  mutate(genome_id = sapply(strsplit(as.character(V1),"\\/"),'[',4)) %>%
  mutate(genome_id = sapply(strsplit(as.character(genome_id),"\\."),'[',1)) %>%
  dplyr::select(-V1) %>%
  dplyr::rename(mlst=V2)


tojoin = mlst %>%
  left_join(ectyper, by = "genome_id") %>%
  left_join(phylogroups, by = "genome_id") %>%
  clean_names()

length(setdiff(metad$genome_id,tojoin$genome_id))
length(intersect(ncbi_short$genome_id,tojoin$genome_id))
length(setdiff(ncbi_short$genome_id,tojoin$genome_id))

ncbi_short = ncbi_short %>% left_join(tojoin, by="genome_id")

n_distinct(ncbi_short$genome_id)
```


### Now dereplicate and annotate K+ and -
```{r}

ncbi_derep = ncbi_short %>%
  distinct(predicted,manual_ktyp,genome_id, bio_project_accession,
           host_common_name, host_gender, host_age, host_group,host_health,
           isolation_source,geographic_group, geographic_location,
           phylogroup, o_type, h_type, mlst, ANI99,group) %>%
  dplyr::group_by(predicted,manual_ktyp,bio_project_accession,
                  host_common_name, host_gender, host_age, host_group,host_health,
                  isolation_source, geographic_group, geographic_location, ANI99,
                  phylogroup,o_type, h_type, mlst,group) %>%
  dplyr::summarise(count=n())

#this gives 23188 entries
nrow(ncbi_derep)

ncbi_derep = ncbi_derep %>%
  mutate(predicted = case_when(manual_ktyp == "K_pos" ~ predicted,
                                    TRUE ~ "K_neg_partial")) 
```


### Fix K type nomenclature and ST numbers for all downstream analysis and Supp table generation
```{r}
ncbi_derep = ncbi_derep %>%
  mutate(mlst = gsub("\\.0","",mlst))

write.xlsx(ncbi_derep, paste0(outputdir,"ncbi_annotated_derep_260112.xlsx"))

```

### Final file to use
```{r}
ncbi_derep = read.xlsx(paste0(outputdir,"ncbi_annotated_derep_260112.xlsx"))
```

## GUTMAGS

### load Ktypr MAG output
```{r}
gutmags <- fread(paste0(inputdir,"commensal_flank_20260115_ktyps.tsv")) %>%
  dplyr::distinct(genome_id, predicted, pred_nr_genes, pred_genes_in_genome,pred_acc_bitscore,
                  pred_is_complete, KpsEDCSMT_nr_genes, KpsEDCSMT_genes_in_genome,KpsEDCSMT_acc_bitscore,
                  KpsEDCSMT_is_complete, KpsFU_nr_genes,KpsFU_genes_in_genome,
                  KpsFU_acc_bitscore,KpsFU_is_complete) 

n_distinct(gutmags$genome_id)
  
  
Kgroups <- readxl::read_xlsx(paste0(inputdir,"subgroups.xlsx"))

gutmags = gutmags %>%
  left_join(Kgroups, by=c("predicted"="new_name")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           TRUE ~ group)) %>%
  mutate(sushisample = sub("^(.*)_MAG.*", "\\1", genome_id)) %>%
  mutate(sushisample= interaction(sushisample, "METAG",sep="_")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           TRUE ~ group)) %>%
  dplyr::ungroup()

metad <- fread(paste0(inputdir,"gutmags_metadata.tsv"))

gutmags = gutmags %>% left_join(metad, by= "sushisample")
 
#process in same way as ncbi 
gutmags = gutmags %>%
  mutate(manual_ktyp = case_when(pred_is_complete == 1 & KpsEDCSMT_is_complete == 1 ~ "K_pos",
                                 pred_is_complete ==0 & KpsEDCSMT_is_complete == 0 ~ "K_neg",
                                 TRUE ~ "K_partial")) %>%
  mutate(predicted = case_when(manual_ktyp == "K_pos" ~ predicted,
                                    TRUE ~ "K_neg_partial")) %>%
  distinct(genome_id, predicted, manual_ktyp, group, age_category_new, body_site, born_method,breastfeeding_duration, c_section_type, disease, family, gender, gestational_age, non_westernized, premature,study_name, sushisample, treatment,country)

# harmonize metadata with ncbi collection categories
gutmags=gutmags %>%  
  mutate(host_age = case_when(age_category_new %in% c("<=1d", "newborn", "<=1y") ~ "Neonate and infants",
                             age_category_new %in% c("<=3y", "<=18y", "<=11y") ~ "Child",
                             age_category_new %in% c("<=30y", "<=50y", "<=85y", "40to75", "adult") ~ "Adult",
                             TRUE ~ "Unknown"),
         host_gender= case_when(gender == "female" ~ "Female",
                                gender == "male" ~ "Male",
                                TRUE ~ "Unknown"),
         geographic_group = case_when(country %in% c("ARG","COL","PER","SLV") ~ "Center and South America",
                                      country %in% c("CHN") ~ "China and South-East Asia",
                                      country %in% c("DEU","DNK","ESP","EST","FIN","GBR","ITA","SWE") ~ "Europe",
                                      country %in% c("FJI") ~ "Oceania",
                                      country %in% c("GNB","MDG") ~ "Africa",
                                      country %in% c("KAZ","RUS") ~ "North and Central Asia" ,
                                      country %in% c("USA") ~ "North America",
                                      is.na(country) ~ "Unknown"),
         host_health = case_when(grepl("ascites|cirrhosis|fatty_liver", disease)  ~ "Liver Disease",
                                     grepl("respiratoryinf", disease)  ~ "Pneumonia and Pleuritis",
                                     grepl("infectiousgastroenteritis", disease)  ~ "GI Tract and Diarrhoea",
                                     grepl("pyelonephritis", disease)  ~ "Urinary Tract Infection",
                                     grepl("migraine", disease)  ~ "Non-Infectious Neurological and Psychiatric Conditions" ,
                                     grepl("generic_diabetes", disease)  ~ "Diabetes",
                                     grepl("asthma", disease)  ~ "asthma",
                                     grepl("premature_born", disease)  ~ "Premature",
                                     grepl("IBD", disease)  ~ "IBD",
                                     grepl("otitis", disease)  ~ "Otitis",
                                     TRUE ~ "Healthy"))


gutmags = gutmags %>%
  dplyr::select(-c(age_category_new,gender,country)) %>%
  # keep only MAGs from healthy individuals
 filter(!(host_health %in% c("IBD", "GI Tract and Diarrhoea","Urinary Tract Infection","Premature")))
  
n_distinct(gutmags$genome_id)
#2763
```

### join ectyper mlst with gutmags
```{r}
# assemble metadata from mlst, ectyper and phylogroups
phylogroups = fread(paste0(inputdir, "assigned_phylogroups_MAGs.csv"))
phylogroups[duplicated(phylogroups$genome_id), ]
n_distinct(phylogroups$genome)
# 3122 genomes
# for one MAG phylogroup couldnt be determined

phylogroups = phylogroups %>%
  mutate(genome_id = sapply(strsplit(as.character(genome),"\\."),'[',1)) %>%
  dplyr::select(-c(genome, distance))

ectyper = fread(paste0(inputdir,"ectyper_ecoli_MAGs.tsv")) %>%
  filter(Name != "Name")
ectyper[duplicated(ectyper$Name), ]
n_distinct(ectyper$Name)
# 3123 genomes
ectyper = ectyper %>%
  mutate(genome_id = sapply(strsplit(as.character(Name),"\\."),'[',1)) %>%
  dplyr::select(c(`O-type`, `H-type`,genome_id)) %>%
  clean_names()

setdiff(ectyper$genome_id,phylogroups$genome_id)
# [1] "BAEC15-1_SAMEA2580062_MAG_00000002"
# this is the one without phylogroup

mlst = fread(paste0(inputdir,"mlst_MAGs.tsv"), sep = "\t", fill=TRUE,header=F)
mlst[duplicated(mlst$V1), ]
n_distinct(mlst$V1)

mlst = mlst %>%
  mutate(genome_id = gsub("^genomes/|\\.fa$", "", V1)) %>%
  dplyr::rename(mlst=V3) %>%
  dplyr::select(c(genome_id,mlst))

setdiff(ectyper$genome_id,mlst$genome_id)
setdiff(phylogroups$genome_id,mlst$genome_id)

gutmags = gutmags %>% left_join(mlst) %>% left_join(ectyper) %>% left_join(phylogroups) %>%
  mutate(mlst = gsub("\\.0","",mlst))

n_distinct(gutmags$genome_id)

# remove phylogroup- MAG
gutmags = gutmags %>%
  filter(!is.na(phylogroup))
```


# ED Table 13 and final file to use

```{r}
# write file to use for downstream analyses
write.xlsx(gutmags, paste0(outputdir,"gutmags_annotated_metad_ectyper.xlsx"), sep ="\t")
# write Table S5
write.xlsx(gutmags, paste0(outputdir,"EDTable_13.xlsx"))
```


## EGLI

### load Ktypr egli output
```{r}
egli <- fread(paste0(inputdir,"egli_flank_20260115_ktyps.tsv"))

Kgroups <- readxl::read_xlsx(paste0(inputdir,"subgroups.xlsx"))

egli = egli %>%
  left_join(Kgroups, by=c("predicted"="new_name")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           TRUE ~ group))

#process in same way as ncbi 
egli = egli %>%
  mutate(manual_ktyp = case_when(pred_is_complete == 1 & KpsEDCSMT_is_complete == 1 ~ "K_pos",
                                 pred_is_complete ==0 & KpsEDCSMT_is_complete == 0 ~ "K_neg",
                                 TRUE ~ "K_partial")) %>%
  mutate(predicted = case_when(manual_ktyp == "K_pos" ~ predicted,
                                    TRUE ~ "K_neg_partial")) %>%
  distinct(genome_id, predicted, manual_ktyp, group)
```


### annotate with metadata
```{r}
txtmetadegli = readxl::read_xlsx(paste0(inputdir,"Egli_UPEC_metadata_fromTim_250307.xlsx"),sheet = 1) %>% 
  clean_names()


egli = egli %>% 
  mutate(seq_id = sapply(strsplit(as.character(genome_id),"_"),'[',2)) %>%
  left_join(txtmetadegli, by = "seq_id")
  

# select the 825 distinct strains from distinct patients
goldenset825 <- fread(paste0(inputdir,"825_strains_included.txt"),header =F) %>%
  mutate(V1 = gsub("-wh*","", V1)) %>%
  mutate(V1 = gsub("-1-","-", V1)) %>%
  mutate(V1 = gsub("-A-","-", V1)) %>%
  mutate(V1 = gsub("-19-1","-19", V1)) %>%
  mutate(V1 = gsub("-19-5","-19", V1)) %>%
  mutate(V1 = gsub("703075-202","703075-20", V1)) %>%
  mutate(V1 = gsub("703294-202","703294-20-", V1)) %>%
  mutate(V1 = gsub("703315-203","703315-20", V1)) %>%
  mutate(V1 = gsub("704060-202","704060-20", V1)) %>%
  mutate(V1 = gsub("712679-192","712679-19", V1)) %>%
  mutate(V1 = gsub("13087410-19","130874-19", V1)) 
  
length(intersect(goldenset825$V1, egli$genome_id_not_unique_for_lookup))

egli = egli %>% 
  filter(genome_id_not_unique_for_lookup %in% goldenset825$V1) %>%
  dplyr::distinct(genome_id,genome_id_not_unique_for_lookup, predicted, manual_ktyp, group, gender,age,invasiveness) %>%
  mutate(host_age = "Adult",
         host_gender = case_when(is.na(gender) ~ "Unknown",
                                 gender == "w" ~ "Female",
                                 gender == "m" ~ "Male"),
        geographic_group = "Europe",
        host_health = case_when(invasiveness == "invasive" ~ "Bloodstream Infection",
                                           TRUE ~ "Urinary Tract Infection")) %>%
  dplyr::select(-gender)
  
n_distinct(egli$genome_id_not_unique_for_lookup)
nrow(egli)

```

### join ectyper mlst with egli
```{r}
## prep egli metadata from mlst, ectyper and phylogroups
phylogroups = fread(paste0(inputdir, "assigned_phylogroups_Egli.csv"))
phylogroups[duplicated(phylogroups$genome_id), ]
n_distinct(phylogroups$genome)
#1072

phylogroups = phylogroups %>%
  mutate(genome_id = sapply(strsplit(as.character(genome),"\\."),'[',1)) %>%
  dplyr::select(-c(genome, distance))

ectyper = fread(paste0(inputdir,"ectyper_egli.tsv")) %>%
  filter(Name != "Name")
ectyper[duplicated(ectyper$Name), ]
n_distinct(ectyper$Name)

ectyper = ectyper %>%
  mutate(genome_id = sapply(strsplit(as.character(Name),"\\."),'[',1)) %>%
  dplyr::select(c(`O-type`, `H-type`,genome_id)) %>%
  clean_names()
#1076

mlst = fread(paste0(inputdir,"egli_genomes_mlst.tsv"), sep = "\t", fill=TRUE,header=F)
mlst[duplicated(mlst$V1), ]
n_distinct(mlst$V1)

mlst = mlst %>%
  mutate(genome_id = gsub(".*/([^/]+)\\.scaffolds\\.min200\\.fasta", "\\1", V1)) %>%
  dplyr::rename(mlst=V3) %>%
  dplyr::select(c(genome_id,mlst))

setdiff(ectyper$genome_id,mlst$genome_id)

# join
egli = egli %>% left_join(mlst) %>% left_join(ectyper) %>% left_join(phylogroups) %>%
  mutate(mlst = gsub("\\.0","",mlst))


```

### fix multiple assignments for same strain isolated multiple times from same patients
```{r}


egli = egli %>%
  mutate(predicted = case_when(genome_id_not_unique_for_lookup %in% c("130874-19") ~ "K5",
                               TRUE ~ predicted),
         manual_ktyp = case_when(genome_id_not_unique_for_lookup %in% c("130874-19") ~ "K_pos",
                               TRUE ~ manual_ktyp),
         group = case_when(genome_id_not_unique_for_lookup %in% c("130874-19") ~ "G2A",
                               TRUE ~ group)) %>%
  select(-genome_id) %>%
  distinct()

nrow(egli)

egli = egli %>% 
  mutate(invasiveness = case_when(host_health == "Urinary Tract Infection" ~ "non-invasive",
                                  TRUE ~ invasiveness),
         isolation_source = case_when(invasiveness == "invasive" ~ "Blood",
                                                 TRUE ~ "Urine and bladder"))

egli = egli %>%
  mutate(host_group = "Human")
```

### Final file to use
```{r}
write.xlsx(egli, paste0(outputdir,"egli_annotated_metad_ectyper.xlsx"), sep ="\t")
```

# ED Table 12

```{r}
egli = readxl::read_xlsx(paste0(outputdir,"egli_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group),
         phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup))

sheets <- list("annotated_NCBI_collection" = ncbi_short, "dereplicated_NCBI_collection" = ncbi_derep, "Cuenod_collection" = egli)

write_xlsx(sheets,paste0(outputdir,"EDTable_12.xlsx"))

rm(sheets)
```


#------------FIGURES-----------

# Fig. 3

## prep data (NCBI + EGLI)


```{r}
ncbi_derep = readxl::read_xlsx(paste0(outputdir,"ncbi_annotated_derep_260112.xlsx"))

### prep data
ncbi_analysis = ncbi_derep %>% 
  mutate(phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup),
         group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
  dplyr::ungroup()

# ADD EGLI
# annotate groups 
egli = readxl::read_xlsx(paste0(outputdir,"egli_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group),
         phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup),
         host_group = "Human",
         host_common_name = "Human",
         geographic_location = "Switzerland",
         isolation_source = case_when(host_health == "Bloodstream Infection" ~ "Blood",
                                      host_health != "Bloodstream Infection" ~ "Urine and bladder"),
         manual_ktyp = case_when(predicted == "K_neg_partial" ~ "K_neg",
                                 TRUE ~ "K_pos"))


common_cols <- intersect(names(ncbi_analysis), names(egli))
totest <- gdata::combine(ncbi_analysis[common_cols], egli[common_cols])
         
```

## write this for Fig. 4, S4-5
```{r}
writexl::write_xlsx(totest, paste0(outputdir, "ncbi_egli_combined_for_analysis.xlsx"))
```


### for text
kps+ fraction
```{r}
totest %>% group_by(manual_ktyp) %>% summarise(count=n(),
                                               proportion = count/nrow(.))

totest %>% group_by(host_group) %>% summarise(count=n(),
                                               proportion = count/nrow(.))

totest %>% group_by(geographic_group) %>% summarise(count=n(),
                                               proportion = count/nrow(.))
```


##Fig. 3B. kps+ fraction
```{r}
toplot = totest %>%
  mutate(posorneg = case_when(manual_ktyp == "K_pos" ~ "K_pos",
                              TRUE ~ "K_neg")) %>%
  # distinguish between pets and wild animals
  mutate(host_group = case_when(host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
                                host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
                                host_group == "Environmental" ~ "Other Environmental Source",
                                host_group == "Human-/animal-\nassociated Environment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
                                TRUE ~ host_group)) %>%
  dplyr::group_by(host_group) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(total, host_group, posorneg) %>%
  dplyr::summarise(count = n()) %>%
  ungroup() %>%
  mutate(proportion = (count/total)*100)

toplot$host_group <- factor(toplot$host_group, levels=rev(c("Human","Human-/animal-\nassociated Environment","Pets","Avian","Livestock","Livestock-\nassociated Environment","Wild Animals","Food","Other Environmental Source","Laboratory","Unknown")))


pdf(paste0(plotdir,"3B_ncbi_kps+.pdf"),width = 8,height=5)
ggplot(toplot,aes(x=host_group,y=proportion,fill=posorneg,label=count)) +
  geom_bar(stat = "identity", color = "black") +
   geom_text(aes(y = proportion),
            position = position_stack(vjust = 0.5), # Center labels within each bar segment
            size = 5, 
            color = "black",
            check_overlap = TRUE) + 
  scale_fill_manual(values =c("white","grey"), label = c("-","+")) +
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=60,hjust=1)) +
  labs(y = "% genomes", x= "", fill = expression(italic("kps"))) +
  coord_flip() +
  theme_classic() +
  theme(axis.text.x = element_text(hjust=0.5, size = 15),
        axis.title=element_text(size=15),
        axis.text.y = element_text(size = 15),
        legend.position="bottom",
        legend.text = element_text(size=10),
        legend.title = element_text(size=15))

dev.off()
```

##Fig. 3C - host capsule group proportion

### for text
```{r}

toplot = totest %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
  # distinguish between pets and wild animals
  mutate(host_group = case_when(host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
                                host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
                                host_group == "Environmental" ~ "Other Environmental Source",
                                host_group == "Human-/animal-associated\nEnvironment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
                                TRUE ~ host_group)) %>%
  filter(manual_ktyp == "K_pos") %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(total, group) %>%
  dplyr::summarise(count = n()) %>%
  ungroup() %>%
  mutate(proportion = (count/total)*100)



```

### figure 
```{r}

mycolors = c(
  "G2A" = "#71B7E2",
  "G2B" = "#345C7B",
  "G3A" = "#7653AB",
  "G3B" = "#A48DCC",
  "kps -" = "grey"
)

toplot = totest %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
  # distinguish between pets and wild animals
  mutate(host_group = case_when(host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
                                host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
                                host_group == "Environmental" ~ "Other Environmental Source",
                                host_group == "Human-/animal-associated\nEnvironment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
                                TRUE ~ host_group)) %>%
  filter(manual_ktyp == "K_pos") %>%
  dplyr::group_by(host_group) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(total, host_group, group) %>%
  dplyr::summarise(count = n()) %>%
  ungroup() %>%
  mutate(proportion = (count/total)*100)


toplot$host_group <- factor(toplot$host_group, levels=rev(c("Human","Human-/animal-\nassociated Environment","Pets","Avian","Livestock","Livestock-\nassociated Environment","Wild Animals","Food","Other Environmental Source","Laboratory","Unknown")))


pdf(paste0(plotdir,"3C_ncbi_group_environments.pdf"),width = 10,height=4.5)
ggplot(toplot,aes(x=host_group,y=proportion,fill=group,label=count)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = mycolors,guide = guide_legend(ncol = 1), na.value = "grey") +
   geom_text(aes(y = proportion),
            position = position_stack(vjust = 0.5), 
            size = 4, 
            color = "white",fontface="bold",
            check_overlap = TRUE) + 
  labs(y = expression(paste("% ", italic("kps"), "+ genomes")), x= "", fill="Capsule Group") +
  coord_flip() +
  theme_classic() +
   theme(axis.text.x = element_text(hjust=0.5, size = 15),
        axis.title=element_text(size=15),
       axis.text.y = element_text(size = 15, lineheight = 0.6),
        legend.position="bottom",
        legend.direction = "horizontal",
        legend.text = element_text(size=10),
        legend.title = element_text(size=15)) +
  guides(fill=guide_legend(nrow=1))


dev.off()

```


## Fig. 3D. Heatmaps K types environments

```{r}
toplot = totest %>%
  filter(manual_ktyp == "K_pos") %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
    mutate(host_group = case_when(host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
                                host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
                                host_group == "Environmental" ~ "Other Environmental Source",
                                host_group == "Human-/animal-\nassociated Environment" & host_common_name %in% c("Pig","Cow","Sheep","Chicken","Other or unspecified livestock") ~ "Livestock-associated Environment",
                                host_group == "Human-/animal-\nassociated Environment" & !(host_common_name %in% c("Pig","Cow","Sheep","Chicken","Other or unspecified livestock")) ~ "Human-/animal-associated Environment",
                                TRUE ~ host_group)) %>%
  dplyr::group_by(host_group) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(total, host_group,predicted) %>%
  dplyr::summarise(count = n()) %>%
  ungroup() %>%
  mutate(proportion = (count/total)*100) %>%
  filter(predicted != "K_neg_partial") %>%
  mutate(host_group=paste0(host_group, " (n=",total,")")) %>%
  complete(host_group, predicted, fill = list(proportion = NA, label = ""))

#--------clustered version--------------

mat_data = toplot %>%
  distinct(predicted,host_group, proportion) %>%
  pivot_wider(names_from = predicted,values_from = proportion)

mat_data[is.na(mat_data)] <- 0

mat_data = mat_data %>% as.data.frame() %>%
  set_rownames(.$host_group) %>%
  dplyr::select(-host_group) %>%
  as.matrix()

#remove all zero rows or column
mat_data <- mat_data[rowSums(mat_data != 0) > 0, colSums(mat_data != 0) > 0]

#  dendrogram 
hc <- hclust(as.dist(1-cor(mat_data, method="spearman")), method="complete")
hr <- dendsort(hclust(as.dist(1-cor(t(mat_data), method="spearman")), method="complete"))

dend1 <- as.dendrogram(hr)
c_group <- 6 

dend2 <- as.dendrogram(hc)

# reannotate with groups 
group_annot = ncbi_derep %>% distinct(predicted,group)
toplot = toplot %>% inner_join(group_annot)

df = data.frame(predicted = colnames(mat_data)) %>%
  left_join(group_annot %>% dplyr::distinct(predicted,group)) %>%
    distinct(predicted, group) %>%
    dplyr::select(-predicted)

anno_df = data.frame(
    group = df$group)


ha = HeatmapAnnotation(df = anno_df,
   annotation_height = unit(4, "mm"),
  col = list(group = c( "G2A" = "#71B7E2","G2B" = "#345C7B","G3A" = "#7653AB","G3B" = "#A48DCC","K_partial_or_neg" = "grey")),
  show_legend = FALSE)



pdf(paste0(plotdir,"3D_clustered_hm_environments_ktypes.pdf"), width=20, height=5)

col_fun <- colorRamp2(c(0, 34), c("white", "grey20"))

hm <- Heatmap(mat_data,
        cluster_rows = dend1,
        cluster_columns = dend2,
        top_annotation = ha,
        row_split = 5, column_split = 10,
        use_raster = TRUE, raster_quality = 5,
        border = TRUE,
        col = col_fun,
        rect_gp = gpar(col = ifelse(mat_data == 0, "grey", NA)),
        heatmap_legend_param = list(title = "% K-type\nin host group"),
        row_title = NULL,
        column_title = NULL,
        row_title_gp = gpar(fontsize = 0),
        column_title_gp = gpar(fontsize = 0),
        row_names_gp = gpar(lineheight = 0.7)
)


draw(hm,
     heatmap_legend_side = "left",
     padding = unit(c(2, 1, 1, 3.5), "cm")) 


dev.off()


```




## Fig. 3E - novel K-types 

### for text

```{r}
toplot = totest %>%
  filter(manual_ktyp == "K_pos") %>%
 mutate(host_group = case_when(host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
                                host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
                                host_group == "Environmental" ~ "Other Environmental Source",
                                host_group == "Human-/animal-\nassociated Environment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
                                host_group == "Human-/animal-\nassociated Environment" & !(host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock")) | host_group == "Human" ~ "Human and related",
                                TRUE ~ host_group)) %>%
  mutate(num = as.numeric(str_extract(predicted, "\\d+")),
         Knovel = ifelse(num >= 104, "novel", "known")) %>%
  dplyr::group_by(host_group) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(Knovel,total, host_group) %>%
  dplyr::summarise(count = n()) %>%
  mutate(proportion = (count/total)*100) %>%
  ungroup() 


```


### figure
```{r}
toplot = totest %>%
  filter(manual_ktyp == "K_pos") %>%
 mutate(host_group = case_when(host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
                                host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
                                host_group == "Environmental" ~ "Other Environmental Source",
                                host_group == "Human-/animal-\nassociated Environment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
                                TRUE ~ host_group)) %>%
  mutate(num = as.numeric(str_extract(predicted, "\\d+")),
         Knovel = ifelse(num >= 104, "novel", "known")) %>%
  dplyr::group_by(host_group) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(Knovel,total, host_group) %>%
  dplyr::summarise(count = n()) %>%
  mutate(proportion = (count/total)*100) %>%
  ungroup()
 

set.seed(69)
ktype_colors <- setNames(colorRampPalette(brewer.pal(8, "Set3"))(n_distinct(toplot$predicted)), unique(toplot$predicted))

ktype_colors["other"] <- "grey"


toplot$host_group <- factor(toplot$host_group, levels=rev(c("Human","Human-/animal-\nassociated Environment","Pets","Avian","Livestock","Livestock-\nassociated Environment","Wild Animals","Food","Other Environmental Source","Laboratory","Unknown")))

pdf(paste0(plotdir,"3E_ncbi_k_type_novel_old_hosts.pdf"),width = 9,height=4)

ggplot(toplot,aes(x=host_group,y=proportion,fill=Knovel,label=count)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("grey80","grey50")) + 
   geom_text(aes(y = proportion),
            position = position_stack(vjust = 0.5),
            size = 5,
            color = "black",
            check_overlap = TRUE) +
  labs(y = expression(paste("% ", italic("kps+"), " genomes")), x= "", fill="K-type") +
  coord_flip() +
 theme_classic() +
  theme(axis.text.x = element_text(hjust=0.5, size = 15),
        axis.title=element_text(size=15),
        axis.text.y = element_text(size = 15, lineheight=0.6),
        legend.position="right",
        legend.text = element_text(size=12),
        legend.title = element_text(size=15))

dev.off()


```



## Fig. 3F. number of hosts * geography for novel or known k-types
```{r}
toplot = totest %>%
  filter(manual_ktyp == "K_pos") %>%
 mutate(host_group = case_when(host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
                                host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
                                host_group == "Environmental" ~ "Other Environmental Source",
                                host_group == "Human-/animal-\nassociated Environment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
                                TRUE ~ host_group),
        geographic_location= case_when(grepl("pplicable|ollected|>=32",geographic_location) ~ "unknown",
                                       TRUE ~ geographic_location),
        geographic_location = tolower(geographic_location)) %>%
  mutate(num = as.numeric(str_extract(predicted, "\\d+")),
         Knovel = ifelse(num >= 104, "novel", "known"),
         host_group = interaction(host_common_name,geographic_location)) %>%
  ungroup() %>%
  # correct by the number of genomes 
  group_by(predicted) %>%
  mutate(no_genomes = n()) %>%
  dplyr::ungroup() %>%
  distinct(Knovel,predicted, host_group, no_genomes) %>%
  dplyr::group_by(Knovel,predicted) %>%
  dplyr::reframe(count = n(),
                   normcount=count/no_genomes) %>%
  distinct(predicted,Knovel, count, normcount)


pdf(paste0(plotdir,"3F_ncbi_Ktype_novel_unknown_nohostsgeo.pdf"),width = 3.7,height=4.1)

counts <- toplot %>%
  group_by(Knovel) %>%
  summarise(n = n())

wilcox_res <- wilcox.test(normcount ~ Knovel, data = toplot)

p_value <- signif(wilcox_res$p.value, digits = 2)

ggplot(toplot, aes(x=Knovel,y=normcount,color=Knovel)) +
  geom_violin() +
  geom_boxplot(alpha=0.5) +
  geom_jitter(stat = "identity", width =0.1, pch=21, size =3) +
  labs(x = "K-types", y="# unique hosts x geographic location /\n# genomes", color="K-type") +
  scale_color_manual(values = c("grey60","grey20")) +
  scale_y_continuous(limits=c(0,1.12), breaks = c(0,0.25,0.5,0.75,1)) +
  geom_text(data = counts, aes(x = Knovel, y = 0.1, label = paste0("n=", n)), vjust = -0.5, size = 5) +
  geom_signif(comparisons = list(unique(toplot$Knovel)), 
              annotation = paste0("p=", p_value),
              y_position = max(toplot$normcount) * 1.05, 
              tip_length = 0.02, 
              color = "black",
              textsize = 5) + 
  theme_bw() +
  theme(legend.position="none",
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 15)) +
  guides(alpha="none")

dev.off()

```




#Fig. 4 


#### prep data (same as Fig. 3)
```{r}
totest = readxl::read_xlsx(paste0(outputdir, "ncbi_egli_combined_for_analysis.xlsx"))
```


### Fig. 4A. phylogroups-capsule groups
```{r}

toplot = totest %>%
   filter(host_group == "Human" & host_common_name == "Human" & !(isolation_source %in% c("Food","Probiotic","Sewage and wastewater","Meat","Plant")))%>% 
  dplyr::group_by(phylogroup) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(group,total, phylogroup) %>%
  dplyr::summarise(count = n()) %>%
  mutate(proportion = (count/total)*100) %>%
  # remove genomes to which no phylogroup could be assigned
  filter(phylogroup != "NA")

toplot %>% group_by(phylogroup) %>% summarise(n = sum(count))



mycolors = c(
  "G2A" = "#71B7E2",
  "G2B" = "#345C7B",
  "G3A" = "#7653AB",
  "G3B" = "#A48DCC",
  "kps -" = "grey"
)


toplot$phylogroup <- factor(toplot$phylogroup, levels=c("A","B1","C","E1","E2","G","B2-1","B2-2","D1","D2","D3","F"))

pdf(paste0(plotdir,"4A_phylogroup_group_percent_includingKneg_onlyhuman.pdf"),width = 8,height=3)
ggplot(toplot,aes(x=phylogroup,y=proportion,fill=group,label=count)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = mycolors,guide = guide_legend(ncol = 1), na.value = "grey") +
   geom_text(aes(y = proportion),
            position = position_stack(vjust = 0.5),
            size = 5, 
            color = "black",
            check_overlap = TRUE) + 
  labs(y = "% genomes", x= "Phylogroup", fill="Capsule Group") +
  theme_classic() +
   theme(axis.text.x = element_text(hjust=0.5, size = 15),
        axis.title=element_text(size=15),
        axis.text.y = element_text(size = 15),
        legend.position="none")
dev.off()


```


##prep data - adding MAGs and encoding invasiveness

```{r}
### prep ncbi data
ncbi_analysis = read.xlsx(paste0(outputdir,"ncbi_annotated_derep_260112.xlsx")) %>% 
  filter(host_group == "Human" & host_common_name == "Human" & !(isolation_source %in% c("Food","Probiotic","Sewage and wastewater","Meat","Plant"))) %>%
  mutate(isolation_source = case_when(!(isolation_source %in% c("Stool","Urine and bladder", "Blood","Brain and CSF")) ~ "Other body sites",
                                      TRUE ~ isolation_source),
         host_health = case_when(host_health %in% c("asymptomatic bacteriuria", "Urinary Tract Infection", "recurrent uti") | isolation_source == "Urine and bladder" ~ "UTI",
                                 host_health %in% c("Bloodstream Infection", "neonatal bacteremia") | isolation_source == "Blood" ~ "BSI",
                                 host_health %in% c("Meningitis") | isolation_source == "Brain and CSF" ~ "Meningitis",
                                 host_health %in% c("GI Tract and Diarrhoea", "HUS and TTP", "STEC/Shigella/EHEC") ~ "InPEC",
                                 host_health == "Healthy" & isolation_source == "Stool" ~ "Healthy (stool)",
                                 host_health == "Healthy" & isolation_source != "Stool"~ "Healthy (other)",
                                 TRUE ~ "Other disease")) %>%
  dplyr::ungroup() %>%
  #add genome id column for merging with other datasets - after dereplication we didn't have it anymore
  mutate(genome_id = row_number(),
        group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
  dplyr::select(genome_id, predicted, host_health,host_age,geographic_group, host_gender, group)

# prep healthy gut MAGs
gutmags = readxl::read_xlsx(paste0(outputdir,"gutmags_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group),
         phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup)) %>%
  dplyr::ungroup() %>%
   dplyr::select(genome_id,predicted, host_health,host_age, host_gender, geographic_group, group) %>%
    mutate(host_health = "MAGs")


# prep egli collection
egli = readxl::read_xlsx(paste0(outputdir,"egli_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group),
         host_health = case_when(invasiveness=="invasive" ~ "BSI",
                                                 TRUE ~ "UTI")) %>%
  dplyr::select(predicted, host_gender, host_health,host_age,geographic_group, group)


common_cols <- intersect(intersect(names(ncbi_analysis), names(egli)), names(gutmags))
totest_invasive <- gdata::combine(ncbi_analysis[common_cols], egli[common_cols], gutmags[common_cols])
```

### Fig. 4B - K-type group proportion in disease

```{r}
test = totest_invasive %>%
  dplyr::group_by(host_health) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(group,total, host_health) %>%
  dplyr::summarise(count = n()) %>%
  mutate(proportion = (count/total)*100) %>%
  ungroup()


mycolors = c(
  "G2A" = "#71B7E2",
  "G2B" = "#345C7B",
  "G3A" = "#7653AB",
  "G3B" = "#A48DCC",
  "kps -" = "grey"
)


test$host_health <- factor(test$host_health, levels=rev(c("MAGs","Healthy (stool)","Healthy (other)","InPEC", "UTI","BSI","Meningitis","Other disease")))


pdf(paste0(plotdir,"4B_top_Kgroup_disease.pdf"), width = 7, height = 3)

totals <- test %>%
  group_by(host_health) %>%
  summarise(total_count = sum(count), .groups = "drop")

ggplot(test, aes(x = host_health, y = proportion, fill = group, label = count)) +
  geom_bar(stat = "identity", color="black") +
  scale_fill_manual(values = mycolors,guide = guide_legend(ncol = 1), na.value = "grey") +
  labs(y = "% genomes", x = "", fill = "K-type") +
  scale_y_continuous(breaks=c(0,25,50,75,100), limits=c(0,110)) +
  geom_text(data = totals, aes(x = host_health, y = 105, label = total_count),
            inherit.aes = FALSE,nudge_y = -3,
            size = 4.5, hjust = 0.1) +
  theme_classic() +
  coord_flip() +
  theme(axis.text.x = element_text(size = 16),
        axis.title = element_text(size = 17),
        axis.text.y = element_text(size = 16),
        strip.text = element_text(size = 17),
        legend.position = "none")


dev.off()



```



### Fig. 4C - PCoA isolation source on human K-types (NCBI + egli + MAGs)
same data as 4B

```{r}
test = totest_invasive %>%
  group_by(host_health) %>%
  dplyr::mutate(total = n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(total, host_health, predicted) %>%
  dplyr::summarise(count = n()) %>%
  ungroup() %>%
  mutate(proportion = (count/total) * 100) %>%
  #filter low frequency types
  filter(proportion > 10) %>%
  dplyr::select(host_health, predicted, proportion) %>%
  ungroup() %>%
  pivot_wider(names_from = host_health, values_from = proportion, values_fill = 0) %>% 
  dplyr::select(-c(predicted)) %>%
  dplyr::filter(if_any(everything(), ~ . != 0))


bray_curtis_mat = vegdist(t(test), method = "bray")

# Principle Coordinate Analysis using the cmdscale function.
bray_curtis_pcoa = cmdscale(bray_curtis_mat, k = 4, eig = T)

# Extract coordinates of samples on PCoA axes 1 and 2 and combine with metadata on CRC/CTR.
pcoa_to_visualize = data.frame(bray_curtis_pcoa$points) %>%
  rownames_to_column(var = "host_health") %>%
  dplyr::rename(PCoA1=X1,
                PCoA2=X2)

pcoa_to_visualize = pcoa_to_visualize %>%
  mutate(host_macrocategory= sapply(strsplit(as.character(host_health),"\\."),'[',1))

# Define colors for unique hosts
num_colors <- length(unique(pcoa_to_visualize$host_health))
palette_colors <- colorRampPalette(brewer.pal(9, "Set2"))(num_colors)

# Visualise 
pdf(paste0(plotdir,"4C_pcoA_isolation_site_human.pdf"),width=3,height=2)

eig_vals <- bray_curtis_pcoa$eig

# keep only positive eigenvalues
eig_vals_pos <- eig_vals[eig_vals > 0]

# percentage variance explained
eig_perc <- eig_vals_pos / sum(eig_vals_pos) * 100


ggplot(pcoa_to_visualize, aes(x = PCoA1, y = PCoA2)) + 
  geom_point(aes(fill = host_health), shape = 21, size = 3, colour = "grey40") + 
  geom_text_repel(aes(label = host_health, color = host_health), fontface = "bold", size = 3) +
  scale_fill_manual(values = c("BSI" = "#c46c77", "Meningitis" = "#facdcd", "UTI" = "#afd98f", 
                               "Healthy (stool)" = "#785b5b", "InPEC" = "#a19762", "MAGs" = "#732b29",
                               "Healthy (other)" = "lightblue", "Other disease" = "grey")) +
  scale_color_manual(values = c("BSI" = "#c46c77", "Meningitis" = "#facdcd", "UTI" = "#afd98f", 
                                "Healthy (stool)" = "#785b5b", "InPEC" = "#a19762", "MAGs" = "#732b29",
                                "Healthy (other)" = "lightblue", "Other disease" = "grey")) +
  theme_bw() + 
  labs(x = paste0("PCoA1 (", round(eig_perc[1], 1), "%)"),
       y = paste0("PCoA2 (", round(eig_perc[2], 1), "%)")) +
  theme(legend.position = "none",
        axis.title=element_text(size = 8))

dev.off()
```


###Fig. 4D - GLMM

same data as for 4B-C but restricting to BSI and meningitis vs healthy stool.

#### 1. data prep
```{r}

# ncbi isolates (subset)
ncbi_analysis = read.xlsx(paste0(outputdir,"ncbi_annotated_derep_260112.xlsx")) %>% 
  filter(host_group=="Human" & host_common_name == "Human" & !(isolation_source %in% c("Food","Probiotic","Sewage and wastewater","Meat","Plant"))) %>%
  filter(isolation_source %in% c("Blood", "Brain and CSF") | host_health %in% c("Bloodstream Infection","Meningitis", "neonatal bacteremia","Healthy")) %>%
  mutate(expec = case_when(host_health != "Healthy" ~ 1,
                           TRUE ~ 0)) %>%
  dplyr::ungroup() %>%
  dplyr::select(predicted, host_gender, host_age, mlst, geographic_group, phylogroup,o_type,h_type, expec)

# add healthy gut mags
gutmags = readxl::read_xlsx(paste0(outputdir,"gutmags_annotated_metad_ectyper.xlsx")) %>%
 filter(!(host_health %in% c("IBD", "GI Tract and Diarrhoea","Urinary Tract Infection","Premature"))) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           TRUE ~ group),
         expec = 0,
         phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup)) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
  dplyr::ungroup() %>%
   dplyr::select(predicted, host_gender, host_age, mlst, geographic_group, phylogroup,o_type,h_type, expec)

# add egli
egli = readxl::read_xlsx(paste0(outputdir,"egli_annotated_metad_ectyper.xlsx")) %>%
  filter(invasiveness == "invasive") %>%
  mutate(expec = 1) %>%
   dplyr::select(predicted, host_gender, host_age, mlst, geographic_group, phylogroup,o_type,h_type, expec)


totest = gdata::combine(ncbi_analysis, gutmags, egli) %>%
  filter(phylogroup != "NA") %>%
   mutate(mlst = gsub("\\.0","",mlst))

# fix O-type annotation as in fig. S6A
#-------- harmonise O types
harmonize_serotype <- function(x) {
  if (grepl("/", x)) {
    parts <- unlist(strsplit(x, "/"))
    parts <- sort(parts)
    return(paste(parts, collapse = "/"))
  } else {
    return(x)
  }
}

totest$o_type <- sapply(totest$o_type, harmonize_serotype)


totest %>% group_by(expec,source) %>%
  summarise(n = n())
```


#### 2. GLMM


```{r}

pacman::p_load(glmmTMB)

# Merge rare factor levels
totest$o_type <- as.factor(ifelse(table(totest$o_type)[totest$o_type] < 15, "Other", totest$o_type))
totest$h_type <- as.factor(ifelse(table(totest$h_type)[totest$h_type] < 10, "Other", totest$h_type))
totest$mlst     <- as.factor(ifelse(table(totest$mlst)[totest$mlst] < 10, "Other", totest$mlst))

# Merge rare geographic groups
totest <- totest %>%
  mutate(geographic_group = case_when(
    geographic_group == "Middle East" ~ "Africa",
    geographic_group == "Japan" ~ "Oceania",
    TRUE ~ geographic_group
  ))


totest = totest %>%
  mutate(host_age = as.factor(host_age),
         geographic_group=as.factor(geographic_group),
         host_gender=as.factor(host_gender))

# pool k-types with 0 in any expected class
predicted_expec_counts <- totest %>%
  group_by(predicted, expec) %>%
  summarise(n = n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = expec, values_from = n, values_fill = 0) %>%
  filter(`0` <= 5 | `1` <= 5) %>%
  pull(predicted)

totest <- totest %>%
  mutate(predicted = ifelse(predicted %in% predicted_expec_counts, "Other", as.character(predicted))) %>%
  mutate(predicted = as.factor(predicted))

# flag K-type at risk (low sample size in combination with other factors)
totest <- totest %>%
  group_by(predicted, o_type, h_type, phylogroup, host_age, host_gender, geographic_group, mlst) %>%
  mutate(combo_n = n()) %>%
  ungroup() %>%
  mutate(rare_combo = combo_n < 6)

predicted_flags <- totest %>%
  group_by(predicted) %>%
  summarise(predicted_at_risk = all(rare_combo)) %>%
  ungroup()

totest <- totest %>%
  left_join(predicted_flags, by = "predicted")


model <- glm(
  expec ~ predicted + o_type + h_type + phylogroup + host_age + host_gender + geographic_group + mlst,
  data = totest)


# View(totest %>%
#   group_by(predicted, expec) %>%
#   summarise(count = n()) %>%
#   pivot_wider(names_from = expec, values_from = count))

```


#### 3. convert to OR
```{r}

coefs <- coef(summary(model))
model_results = coefs %>% 
  as.data.frame() %>%
  rownames_to_column(var="term") %>%
  clean_names() %>%
  dplyr::rename(p.value=pr_t) %>%
  mutate(fixed_effect = case_when(grepl("predicted", term) ~ "K-type",
                                  grepl("o_type", term) ~ "O-type",
                                  grepl("phylogroup", term) ~ "Phylogroup",
                                  grepl("mlst", term) ~ "MLST",
                                  TRUE ~ "Other"),
         term_update = gsub("o_type|mlst|phylogroup|predicted", "", term),
         conf.low = estimate - 1.96 * std_error,
         conf.high = estimate + 1.96 * std_error,
         OR = exp(estimate),
         OR_lower = exp(conf.low),
         OR_upper = exp(conf.high),
         significance=case_when(p.value <= 0.01 ~"0.01",
                                p.value <= 0.05 & p.value > 0.01 ~"0.05",
                                TRUE ~ "Not Significant"))

write.table(model_results, paste0(outputdir,"OddsRatio_GLMM_260122.tsv"), sep = "\t",quote = F,row.names = F)
```


### Table S6

```{r}
towrite = model_results %>%
  filter(fixed_effect == "K-type") %>%
  dplyr::select(term_update, std_error,p.value,OR,OR_lower,OR_upper) %>%
  dplyr::rename(term=term_update)
  

write.xlsx(towrite, paste0(outputdir,"EDTable_14.xlsx"))

```


### Figure
```{r}

pacman::p_load(ggbreak)

test = fread(paste0(outputdir,"OddsRatio_GLMM_260122.tsv"), sep = "\t")


pdf(paste0(plotdir,"4D_OR_glmer_predicted_expec_forfigure.pdf"), width = 8, height = 3)

ggplot(test %>% 
         filter(fixed_effect %in% c("K-type")), 
         #filter(fixed_effect %in% c("K-type")), 
       aes(y = reorder(term_update, -OR), x = OR, xmin = OR_lower, xmax = OR_upper, color = significance)) +
  geom_errorbar(size = 0.5, width = 0.3) +
  geom_point() +
  scale_color_manual(
    values = c("0.01" = "darkblue", "0.05" = "dodgerblue2", "Not Significant" = "gray"),
    labels = c("Not Significant" = "p ≥ 0.05", "0.05" = "p < 0.05", "0.01" = "p < 0.01")
  ) +
  theme_bw() +
 
  geom_vline(xintercept = 1, #linetype = "11",
             color = "black", linewidth = 0.1) +
  labs(x = "Odds Ratio", y = "K-types", color = "") +
  coord_flip() +
  theme(
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle = 60, hjust = 1),
    strip.text = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  )


dev.off()

```



# ---------- SUPPLEMENTARY FIGS------------

## ED Fig. 3

```{r}

refseq_short = fread(paste0(inputdir,"refseq37_flank_20260107_ktyps.tsv")) %>%
  dplyr::distinct(genome_id, predicted, pred_nr_genes, pred_genes_in_genome,pred_acc_bitscore,pred_is_complete, KpsEDCSMT_nr_genes,
                  KpsEDCSMT_genes_in_genome,
                  KpsEDCSMT_acc_bitscore, KpsEDCSMT_is_complete,
                  KpsFU_nr_genes,KpsFU_genes_in_genome,
                  KpsFU_acc_bitscore,KpsFU_is_complete) %>%
  mutate(manual_ktyp = case_when(pred_is_complete == 1 & KpsEDCSMT_is_complete == 1 ~ "K_pos",
                                 pred_is_complete ==0 & KpsEDCSMT_is_complete == 0 ~ "K_neg",
                                 TRUE ~ "K_partial")) 

# quick check of KpsC unidentified  
table(refseq_short$manual_ktyp, refseq_short$predicted)

table(refseq_short$manual_ktyp, refseq_short$pred_is_complete, refseq_short$KpsEDCSMT_is_complete)

```

### ED3A - upset plot categories refseq

```{r}

pacman::p_load(dplyr, UpSetR)

toplot_upset <- refseq_short %>%
  transmute(
    `all K-type-specific genes` = as.integer(pred_is_complete == 1),
    `complete kpsEDCMST`        = as.integer(KpsEDCSMT_is_complete == 1),
    `complete kpsFU`            = as.integer(KpsFU_is_complete == 1),
    kpsC     = as.integer(!predicted %in% c("No KpsC identified")),
    assigned = as.integer(!predicted %in% c("No KpsC identified", "No K-type assigned"))
  ) %>%
  mutate(
    `none` = as.integer(
      `all K-type-specific genes` == 0 &
      `complete kpsEDCMST` == 0 &
      `complete kpsFU` == 0 &
      kpsC == 0 &
      assigned == 0
    )
  )

set_order <- c(
  "all K-type-specific genes",
  "complete kpsEDCMST",
  "complete kpsFU",
  "kpsC",
  "assigned",
  "none"# bottom
)

# Write one PDF per page
base_pattern <- file.path(plotdir, "ED3A_upsetRefseq_%03d.pdf")
final_pdf    <- file.path(plotdir, "ED3A_upsetRefseq.pdf")

pdf(base_pattern, width = 5, height = 2.5, onefile = FALSE)
UpSetR::upset(
  toplot_upset,
  sets = set_order,
  nsets = length(set_order),
  keep.order = TRUE,
  order.by = "freq",
  nintersects = NA,
  
  mb.ratio = c(0.3, 0.3),

  point.size = 1.8,
  line.size  = 0.6,
  text.scale = c(1, 1, 0.85, 0.75, 0.75, 0.85)
)


dev.off()

# Keep page 2 (the plot) and delete page 1 (blank)
file.rename(sprintf(base_pattern, 2), final_pdf)
unlink(sprintf(base_pattern, 1))


```



### for text

```{r}
refseq_short %>%
  #filter(predicted %in% c("No KpsC identified", "No K-type assigned")) %>%
  mutate(assigned = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ "not assigned",
                              TRUE ~ "assigned"),
         kpsC = case_when(predicted %in% c("No KpsC identified") ~ "-",
                              TRUE ~ "+")) %>%
  group_by(manual_ktyp, assigned,kpsC, pred_is_complete,KpsEDCSMT_is_complete, KpsFU_is_complete) %>%
  summarise(n = n()) %>% 
  arrange(assigned)
  

```

### ED3B - barplot K-types

```{r}

Kgroups <- readxl::read_xlsx(paste0(inputdir,"subgroups.xlsx"))

refseq_short = refseq_short %>%
  left_join(Kgroups, by=c("predicted"="new_name")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           TRUE ~ group))

toplot <- refseq_short %>%
  filter(manual_ktyp == "K_pos") %>%
  dplyr::group_by(predicted, group) %>%
  dplyr::mutate(count = n()) %>%
  mutate(proportion = (count / nrow(.)) * 100) %>%
  ungroup() %>%
  dplyr::group_by(predicted, group) %>%
  dplyr::summarise(count = n(), .groups = "drop") %>%
  mutate(proportion = (count / sum(count)) * 100) %>%
  ungroup()


pdf(paste0(plotdir,"ED3B_Refseq_Ktype_prevalence.pdf"),width = 18,height=5)
ggplot(toplot, aes(x = reorder(predicted, -proportion),
                   y = proportion, fill = group)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("G2A" = "#71B7E2",
                               "G2B" = "#345C7B",
                               "G3A" = "#7653AB",
                               "G3B" = "#A48DCC",
                               "K_partial_or_neg" = "grey"),
  guide = guide_legend(ncol = 1),
  na.value = "grey") +
 # coord_flip() +
  geom_text(aes(y = proportion + 1, label = count,angle=90, hjust=0.2),
            size = 5, #angle = 90,
            color = "black", #angle = 90, hjust = -0.1,
            check_overlap = TRUE) + 
  labs(y = expression(paste("% ", italic("kps"), "+ genomes")),
       x= "", 
       fill="Capsule Group", title="") + 
  scale_y_continuous(limits = c(0, 20)) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 15, angle=60, hjust=1),
        axis.title = element_text(size = 18),
        axis.text.y = element_text(size = 15),
        legend.position = "none",
        strip.text = element_text(size = 15))


dev.off()



```


## ED Fig. 6

#### prep data (same as Fig. 3)
```{r}
totest = readxl::read_xlsx(paste0(outputdir, "ncbi_egli_combined_for_analysis.xlsx"))
```


### ED6A. Geography capsule positivity

```{r}
toplot = totest %>%
  mutate(posorneg = case_when(manual_ktyp == "K_pos" ~ "K_pos",
                              TRUE ~ "K_neg")) %>%
  mutate(geographic_group= replace_na(geographic_group, "Unknown")) %>%
  mutate(geographic_group= gsub("^$", "Unknown", geographic_group)) %>%
  dplyr::group_by(geographic_group) %>%
  mutate(total=n()) %>% ungroup() %>%
  dplyr::group_by(geographic_group, total, posorneg) %>%
  summarise(count=n()) %>%
  ungroup() 


pdf(paste0(plotdir,"ED6A_ncbi_kps+_geography.pdf"),width = 8,height=5)
ggplot(toplot,aes(x=reorder(geographic_group, count),y=count,fill=posorneg,label=count)) +
  geom_bar(stat = "identity", color = "black") +
   geom_text(aes(y = total+400, label =total),
            size = 5,
            color = "black",
            check_overlap = TRUE) +
  scale_fill_manual(values =c("white","grey"), label = c("-","+")) +
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=60,hjust=1)) +
  labs(y = "# genomes", x= "", fill="Capsule") +
  coord_flip() +
  theme_classic() +
  theme(axis.text.x = element_text(hjust=0.5, size =15),
        axis.text.y = element_text(size =15),
        legend.position="inside",
        legend.text = element_text(size=8),
        axis.title=element_text(size=15))

dev.off()

```

### ED6B -  Barplot of k-types

##### for text: how many distinct k-types per environment?

```{r}

toplot = totest %>%
  filter(manual_ktyp == "K_pos") %>%
  dplyr::mutate(total = nrow(.)) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(total, predicted) %>%
  dplyr::summarise(count = n()) %>%
  ungroup() %>%
  mutate(proportion = (count/total) * 100)

```



##### figure
```{r}

toplot = totest %>%
  filter(manual_ktyp == "K_pos") %>%
  mutate(host_common_name = case_when(
    host_group == "Environmental" & isolation_source %in% c("Sewage and wastewater", "River", "Seawater", "Sediment and soil","Air and dust","Manure","Unspecified Clinical Isolate") ~ isolation_source,
    host_group == "Environmental" & host_common_name %in% c("Water") & !(isolation_source %in% c("Sewage and wastewater", "River", "Seawater")) ~ "Water",
    host_group == "Human" ~ "Human",
    host_group == "Laboratory" ~ "Laboratory",
    TRUE ~ host_common_name)) %>%
  mutate(host_common_name = gsub("Unmapped", "Unknown", host_common_name),
         host_common_name = gsub("Environmental Source", "Unknown", host_common_name),
         host_common_name = gsub("Food", "Unknown", host_common_name)) %>%
  mutate(host_group = case_when(
    host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
    host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
    host_group == "Environmental" ~ "Other Environmental Source",
    host_group == "Human-/animal-\nassociated Environment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
    TRUE ~ host_group)) %>%
  dplyr::group_by(host_group) %>%
  dplyr::mutate(total = n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(total, host_group, predicted) %>%
  dplyr::summarise(count = n()) %>%
  ungroup() %>%
  mutate(proportion = (count/total) * 100) %>%
  dplyr::select(host_group, predicted, proportion,total)


# --- GROUP LOW PROPORTION K-TYPES (< 0.5%) AS "OTHER" ---
test <- toplot %>%
  mutate(predicted = ifelse(proportion < 8, "Other", predicted)) %>% 
  # Recalculate total proportions
  group_by(host_group, predicted) %>%
  summarise(proportion = sum(proportion), .groups = "drop")  


test$host_group <- factor(test$host_group, levels=rev(c("Human","Human-/animal-\nassociated Environment","Pets","Avian","Livestock","Food","Livestock-\nassociated Environment","Wild Animals","Other Environmental Source","Laboratory","Unknown")))


#take color mapping from ED6E --> need to plot this first if rerunning all figures
color_mapping <- readRDS(paste0(outputdir,"Ktype_color_mapping.rds"))


pdf(paste0(plotdir,"ED6B_ktype_hostdistribution.pdf"),width = 7, height =4.5)

# set order
test <- test %>%
  arrange(host_group, desc(proportion))  
test$predicted <- factor(test$predicted, levels = unique(test$predicted))


# calculate total counts per bar (disease + phylogroup)

totals <- toplot %>%
  group_by(host_group) %>%
  summarise(total_count = unique(total), .groups = "drop")


ggplot(test, aes(x = host_group, y = proportion, fill = predicted)) +
  geom_bar(stat = "identity", color = "black") +  # Dodge for clustered bars
  scale_fill_manual(values = color_mapping,
                  guide = guide_legend(nrow = 3)) +
  labs(title = "",
       x = "", 
       y = expression(paste("% ", italic("kps"), "+ genomes")),
       fill = "K-Type") +
    geom_text(data = totals, aes(x = host_group, y = 105, label = total_count),
            inherit.aes = FALSE,nudge_y = -3,
            size = 3, hjust = 0.2) +
  coord_flip() +
  theme_classic() +
  theme(legend.text= element_text(size=10),
        legend.key.height = unit(3, "mm"),
        legend.key.width = unit(5, "mm"),
        legend.position="bottom",
        axis.text = element_text(size = 11))

dev.off()

```


### ED6C. Heatmap Ktypes individual hosts - CORRELATION VERSION
```{r}
toplot = totest %>%
  filter(manual_ktyp == "K_pos") %>%
  mutate(host_common_name = case_when(host_group == "Environmental" & isolation_source %in% c("Sewage and wastewater", "River", "Seawater", "Sediment and soil","Air and dust","Manure","Unspecified Clinical Isolate") ~ isolation_source,
                                      host_group == "Environmental" & host_common_name %in% c("Water") & !(isolation_source %in% c("Sewage and wastewater", "River", "Seawater")) ~ "Water",
                                      host_group == "Human" ~ "Human",
                                      host_group == "Laboratory" ~ "Laboratory",
                                      TRUE ~ host_common_name)) %>%
    mutate(host_common_name= gsub("Unmapped", "Unknown", host_common_name)) %>%
  mutate(host_common_name= gsub("Environmental Source", "Unknown", host_common_name)) %>%
  mutate(host_common_name= gsub("Food", "Unknown", host_common_name)) %>%
    mutate(host_group = case_when(host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
                                host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
                                host_group == "Environmental" ~ "Other Environmental Source",
                                host_group == "Human-/animal-\nassociated Environment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
                                TRUE ~ host_group)) %>%
  mutate(host_group_name=interaction(host_group,host_common_name, geographic_group)) %>%
  dplyr::group_by(host_group, host_group_name) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  # keep only hosts with > 15
  filter(total >= 15 ) %>%
  dplyr::group_by(total, host_group,host_group_name,predicted) %>%
  dplyr::summarise(count = n()) %>%
  ungroup() %>%
  mutate(proportion = (count/total)*100) %>%
  filter(predicted != "K_neg_partial") %>%
 # filter(proportion > 10) %>%
  mutate(host_group_name=paste0(host_group_name, " (n=",total,")")) %>%
  complete(host_group_name, predicted, fill = list(proportion = NA, label = ""))
  

mat_data = toplot %>%
  distinct(predicted,host_group_name, proportion) %>%
  pivot_wider(names_from = host_group_name,values_from = proportion)

mat_data[is.na(mat_data)] <- 0

df_corr = mat_data %>%
  pivot_longer(cols = c(2:ncol(.)), values_to = "proportion",names_to = "host_group_name") 


df_wide <- df_corr %>%
  pivot_wider(names_from = host_group_name, values_from = proportion)

# calculate correlation between hosts
corr_matrix <- cor(df_wide[, -1], use = "complete.obs", method = "pearson")
corr_long <- as.data.frame(as.table(corr_matrix))

colnames(corr_long) <- c("host1", "host2", "corr")

is.na.data.frame <- function(x)
  do.call(cbind, lapply(x, is.na))

fix_na <- function (x) {
  x[is.na(x)] <- 0
  return(x)
}
corr_long = fix_na(corr_long)

dim(corr_long)


corr_long = corr_long %>%
  as.data.frame() %>%
  pivot_wider(names_from = host2, values_from = corr) %>%
  fix_na() %>%
  as.data.frame() %>%
  set_rownames(.$host1) %>%
  dplyr::select(-host1) %>%
  as.matrix()

dim(corr_long)


# Extract host group and geographical group
row_annotation_data <- data.frame(host_group_name = rownames(corr_long)) %>%
  mutate(host_group = sapply(strsplit(as.character(host_group_name), "\\."), '[', 1),
    geographic_group = sub("^[^.]+\\.[^.]+\\.([^\\(]+) \\(n=.*", "\\1", host_group_name)) %>%
  dplyr::select(-host_group_name)

# Unique values for coloring
unique_host_groups <- unique(row_annotation_data$host_group)
unique_geo_groups <- unique(row_annotation_data$geographic_group)

# Generate colors for annotations
set.seed(12)
host_group_colors <- setNames(colorRampPalette(brewer.pal(8, "Set3"))(length(unique_host_groups)), unique_host_groups)
geo_group_colors <- setNames(colorRampPalette(brewer.pal(8, "Paired"))(length(unique_geo_groups)), unique_geo_groups)



# Create the row annotation
ra <- rowAnnotation(
  hostgroup = row_annotation_data$host_group,
   Geographical_Group = row_annotation_data$geographic_group,
  annotation_label = c("Host", "Geography"),
  col = list(hostgroup = host_group_colors,
             Geographical_Group = geo_group_colors),
  annotation_width = unit(4, "mm")
)

rownames(corr_long) <- sub("^[^.]+\\.", "", rownames(corr_long))


library(circlize)
col_fun = colorRamp2(c(min(corr_long), 0, max(corr_long)), hcl_palette = "RdBu")
col_fun(seq(-20, 20))

# Increase space between heatmap body and the heatmap legend (right side)
ht_opt$HEATMAP_LEGEND_PADDING = unit(10, "mm")   # try 5–20 mm

# If you also have annotation legends on the right:
ht_opt$ANNOTATION_LEGEND_PADDING = unit(10, "mm")


ht <- Heatmap(
  corr_long,
  row_names_gp = gpar(fontsize = 9),
  column_title="host-location",
  row_title ="host-location",
  show_column_names = FALSE,
  col=col_fun,
  row_split = 10,
  column_split = 10,
  heatmap_legend_param = list(
    title = "Pearson correlation coefficient",
    legend_direction = "horizontal"
  ),
  left_annotation = ra,
  use_raster = TRUE, raster_quality = 5
)

pdf(paste0(plotdir,"ED6C_hm_hosts_corr_matrix.pdf"), width=14, height=10)

draw(
  ht,
  heatmap_legend_side = "right",
)



dev.off()

ht_opt(RESET = TRUE)
```


### ED6D. K-type frequency overall
```{r}

toplot <- totest %>%
  filter(manual_ktyp == "K_pos") %>%
  dplyr::group_by(predicted, group) %>%
  dplyr::mutate(count = n()) %>%
  mutate(proportion = (count / nrow(.)) * 100) %>%
  ungroup() %>%
  mutate(predicted = case_when(proportion < 0.5 ~ "other",
                               TRUE ~ predicted)) %>%
  dplyr::group_by(predicted, group) %>%
  dplyr::summarise(count = n(), .groups = "drop") %>%
  mutate(proportion = (count / sum(count)) * 100) %>%
  ungroup()




pdf(paste0(plotdir,"ED6D_ncbi_Ktype_prevalence.pdf"),width = 6,height=10)
ggplot(toplot, aes(y = factor(predicted, levels = c("other",setdiff(predicted[order(proportion)], "other"))),
                   x = proportion, fill = group)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = mycolors, guide = guide_legend(ncol = 1), na.value = "grey") +
 # coord_flip() +
  geom_text(aes(x = proportion + 1, label = count),
            size = 5, #angle = 90,
            color = "black", #angle = 90, hjust = -0.1,
            check_overlap = TRUE) + 
  labs(x = expression(paste("% ", italic("kps"), "+ genomes")),
       y= "", 
       fill="Capsule Group", title="") + 
  scale_x_continuous(limits = c(0, 20)) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 15),
        axis.title = element_text(size = 18),
        axis.text.y = element_text(size = 15),
        legend.position = "none",
        strip.text = element_text(size = 15))


dev.off()

```

##### for text
```{r}

toplot <- totest %>%
   mutate(host_common_name = case_when(
    host_group == "Environmental" & isolation_source %in% c("Sewage and wastewater", "River", "Seawater", "Sediment and soil","Air and dust","Manure","Unspecified Clinical Isolate") ~ isolation_source,
    host_group == "Environmental" & host_common_name %in% c("Water") & !(isolation_source %in% c("Sewage and wastewater", "River", "Seawater")) ~ "Water",
    host_group == "Human" ~ "Human",
    host_group == "Laboratory" ~ "Laboratory",
    TRUE ~ host_common_name)) %>%
  mutate(host_common_name = gsub("Unmapped", "Unknown", host_common_name),
         host_common_name = gsub("Environmental Source", "Unknown", host_common_name),
         host_common_name = gsub("Food", "Unknown", host_common_name)) %>%
  mutate(host_group = case_when(
    host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
    host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
    host_group == "Environmental" ~ "Other Environmental Source",
    host_group == "Human-/animal-\nassociated Environment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
    TRUE ~ host_group)) %>%
  filter(manual_ktyp == "K_pos") %>%
  dplyr::group_by(predicted, host_group) %>%
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::group_by(host_group) %>% 
  mutate(proportion = (count / sum(count)) * 100) %>%
  ungroup()
```


### ED6E. geography

```{r}
toplot <- totest %>%
  filter(manual_ktyp == "K_pos") %>%
mutate(host_group = case_when(
    host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
    host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
    host_group == "Environmental" ~ "Other Environmental Source",
    host_group == "Human-/animal-\nassociated Environment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
    TRUE ~ host_group)) %>%
  dplyr::group_by(host_group, geographic_group, predicted) %>%
  dplyr::summarise(count = n(), .groups = "drop") %>%  # Count each K-type within the group
  dplyr::group_by(host_group, geographic_group) %>%
  dplyr::mutate(proportion = (count / sum(count)) * 100) %>%  # Normalize within group
  ungroup() %>%
  filter(!(host_group %in% c("Laboratory","Unknown")) & !(geographic_group %in% c("Asia", "Atlantic Ocean")))



# this figure has the highest diversity of k-types so unified K-type legend should be generated here
# group low proportion K-types as "other"
test <- toplot %>%
  mutate(predicted = ifelse(proportion < 8, "Other", predicted)) %>% 
  # Recalculate total proportions
  group_by(host_group, geographic_group, predicted) %>%
  summarise(proportion = sum(proportion), .groups = "drop")  


# Generate uniform palette to reuse throughout
unique_ktypes <- unique(test$predicted)
unique_ktypes <- unique_ktypes[unique_ktypes != "Other"]
num_colors <- length(unique_ktypes)
palette_colors <- colorRampPalette(brewer.pal(12, "Set3"))(num_colors)
color_mapping <- c(setNames(palette_colors, unique_ktypes), Other = "#B0B0B0")

saveRDS(color_mapping, file = paste0(outputdir,"Ktype_color_mapping.rds"))


# Order legend with Other last
k_levels <- setdiff(names(color_mapping), "Other")
k_levels <- k_levels[order(as.integer(sub("^K", "", k_levels)))]
test$predicted <- factor(test$predicted, levels = c(k_levels, "Other"))


# order for geographic_group
geo_order <- rev(c("Center and South America","North America",  "Europe","Middle East","Africa", "Indian Subcontinent", "China and South-East Asia", "North and Central Asia", "Japan",  "Oceania","Unknown"))
test$geographic_group <- factor(test$geographic_group, levels = geo_order)

# order for host_group
host_order <- c("Human", "Human-/animal-\nassociated Environment", "Pets", "Livestock","Food",  "Livestock-\nassociated Environment", "Avian" ,"Wild Animals", "Other Environmental Source" )
test$host_group <- factor(test$host_group, levels = host_order)

test$predicted <- factor(test$predicted, levels = unique(test$predicted))


pdf(paste0(plotdir,"ED6E_ktype_geography_distribution.pdf"),width = 15, height =7)

totals <- toplot %>%
  group_by(geographic_group, host_group) %>%
  summarise(total_count = sum(count), .groups = "drop")


ggplot(test, aes(x = geographic_group, y = proportion, fill = predicted)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = color_mapping) +  
  geom_text(data = totals, aes(x = geographic_group, y = 105, label = total_count),
            inherit.aes = FALSE,nudge_y = -3,
            size = 3, hjust = 0.2) +
  labs(title = "",
       x = "", 
       y = expression(paste("% ", italic("kps+"), "genomes")),
       fill = "K-Type") +
  coord_flip() +
  scale_y_continuous(limits = c(0,110),breaks=c(0,25,50,75,100)) +
  facet_wrap(~host_group, scales="free") +
  theme_classic() +
  theme(legend.text= element_text(size=8), 
        legend.key.height = unit(4, "mm"),
        legend.key.width = unit(2, "mm"),
        legend.position="right") +
  guides(fill = guide_legend(nrow = 25))


dev.off()


```

##### for text: top K-types per environment

```{r}

toplot <- totest %>%
  filter(manual_ktyp == "K_pos") %>%
mutate(host_group = case_when(
    host_group == "Other Animals" & host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit") ~ "Pets",
    host_group == "Other Animals" & !(host_common_name %in% c("Cat", "Dog", "Horse","Mouse","Rabbit")) ~ "Wild Animals",
    host_group == "Environmental" ~ "Other Environmental Source",
    host_group == "Human-/animal-\nassociated Environment" & host_common_name %in% c("Pig","Cow","Sheep","Other or unspecified livestock") ~ "Livestock-\nassociated Environment",
    TRUE ~ host_group)) %>%
  dplyr::group_by(host_group, predicted) %>%
  dplyr::summarise(count = n(), .groups = "drop") %>%  
  dplyr::group_by(host_group) %>%
  dplyr::mutate(proportion = (count / sum(count)) * 100) %>%
  ungroup() 

```



## ED Fig. 7

### prep data (same as Fig. 3)
```{r}
totest = read_xlsx(paste0(outputdir,"ncbi_egli_combined_for_analysis.xlsx"))     

```

#### ED7A - phylogroups all hosts
```{r}

toplot = totest %>%
  mutate(phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup)) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
  dplyr::group_by(phylogroup) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(group,total, phylogroup) %>%
  dplyr::summarise(count = n()) %>%
  mutate(proportion = (count/total)*100)


mycolors = c(
  "G2A" = "#71B7E2",
  "G2B" = "#345C7B",
  "G3A" = "#7653AB",
  "G3B" = "#A48DCC",
  "kps -" = "grey"
)

toplot = toplot %>%
  filter(phylogroup != "NA")

toplot$phylogroup <- factor(toplot$phylogroup, levels=c("A","B1","C","E1","E2","G","B2-1","B2-2","D1","D2","D3","F"))

pdf(paste0(plotdir,"ED7A_phylogroup_group_percent_includingKneg_allhosts.pdf"),width = 8,height=3)
ggplot(toplot,aes(x=phylogroup,y=proportion,fill=group,label=count)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = mycolors,guide = guide_legend(ncol = 1), na.value = "grey") +
   geom_text(aes(y = proportion),
            position = position_stack(vjust = 0.5), 
            size = 5, 
            color = "black",
            check_overlap = TRUE) + 
  labs(y = "% genomes", x= "Phylogroup", fill="Capsule Group") +
  #coord_flip() +
  theme_classic() +
   theme(axis.text.x = element_text(hjust=0.5, size = 15),
        axis.title=element_text(size=15),
        axis.text.y = element_text(size = 15),
        legend.position="none")
dev.off()


```

##### text 

```{r}

toplot = totest %>%
  mutate(nastyphy = case_when(phylogroup %in% c("B2-1","B2-2","D1","D2","D3","F") ~ "nasty",
                              TRUE ~ "ok")) %>%
  mutate(phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup)) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
  dplyr::group_by(nastyphy) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(group,total, nastyphy) %>%
  dplyr::summarise(count = n()) %>%
  mutate(proportion = (count/total)*100)
```


#### ED7B - isolation source ncbi + egli only Human
```{r}

toplot = totest %>%
   filter(host_group == "Human" & host_common_name == "Human" & !(isolation_source %in% c("Food","Probiotic","Sewage and wastewater","Meat","Plant")))%>% 
  mutate(isolation_source = case_when(isolation_source %in% c("Blood", "Brain and CSF","Urine and bladder","Stool","Unspecified Clinical Isolate") ~ isolation_source,
                                                  TRUE ~ "Other")) %>%
 dplyr::group_by(phylogroup) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(total, phylogroup,isolation_source) %>%
  dplyr::summarise(count = n()) %>%
  mutate(proportion = (count/total)*100)

toplot %>% group_by(phylogroup) %>% summarise(n = sum(count))


toplot = toplot %>%
  filter(phylogroup != "NA")

toplot$phylogroup <- factor(toplot$phylogroup, levels=c("A","B1","C","E1","E2","G","B2-1","B2-2","D1","D2","D3","F"))

toplot = toplot %>%
  mutate(isolation_source = case_when(isolation_source == "Urine and bladder" ~ "Urine and\nbladder",
                                                 isolation_source == "Unspecified Clinical Isolate" ~ "Unspecified Clinical\nIsolate",
                                                 TRUE ~ isolation_source))

pdf(paste0(plotdir,"ED7B_phylogroup_percent_includingKneg_isolationsources_stacked.pdf"),width = 7,height=3.5)
ggplot(toplot , aes(x = phylogroup, y = proportion, fill = isolation_source)) +
 geom_bar(stat = "identity", color = "black") +
   geom_text(aes(y = proportion, label=count),
            position = position_stack(vjust = 0.5),
            size = 4,
            color = "black",
            check_overlap = TRUE) +
  labs(y = "# genomes", x = "", fill = "Isolate source") +
  scale_fill_manual(values = c("Blood" = "#c46c77", "Brain and CSF" = "white", "Urine and\nbladder" = "#cccc99", "Stool" = "#f7ddeb",  "Unspecified Clinical\nIsolate" = "lightblue", "Other" = "grey")) +
  theme_classic() +
  theme(#axis.text.x = element_text(hjust = 1, size = 15,angle=60),
        axis.title = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        axis.text.x = element_text(size = 15),
        strip.text = element_text(size = 15),
        legend.text=element_text(size = 12),
        legend.title=element_text(size = 13),
        legend.position="bottom",
        legend.margin = margin(t = -20, r = -10, b = 1, l = -10),
        legend.direction="horizontal")+
  guides(fill = guide_legend(nrow = 2))
dev.off()


```

#### ED7C - gut commensal global distribution
```{r}
gutmags = readxl::read_xlsx(paste0(outputdir,"gutmags_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           TRUE ~ group),
         expec = 0,
         phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup)) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
  dplyr::ungroup()


toplot = gutmags %>%
  mutate(geographic_group= replace_na(geographic_group, "Unknown")) %>%
  mutate(geographic_group= gsub("^$", "Unknown", geographic_group)) %>%
  dplyr::group_by(geographic_group) %>%
  mutate(total=n()) %>%
  ungroup() %>%
  mutate(posorneg = case_when(manual_ktyp == "K_pos" ~ "K_pos",
                              TRUE ~ "K_neg")) %>%
  dplyr::group_by(geographic_group, total, posorneg) %>%
  summarise(count=n()) %>%
  ungroup() 


pdf(paste0(plotdir,"ED7C_gutmags_kps+_geography.pdf"),width = 7,height=3.5)
ggplot(toplot,aes(x=reorder(geographic_group, count),y=count,fill=posorneg,label=count)) +
  geom_bar(stat = "identity", color = "black") +
   geom_text(aes(y = total+150, label =total),
            size = 5,
            color = "black",#fontface="bold",
            check_overlap = TRUE) +
  scale_fill_manual(values =c("white","grey"), label = c("-","+")) +
  scale_y_continuous(limits=c(0,1780)) +
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=60,hjust=1)) +
  labs(y = "# genomes", x= "", fill="Capsule") +
  coord_flip() +
  theme_classic() +
  theme(axis.text.x = element_text(hjust=0.5, size =15),
        axis.text.y = element_text(size =15),
        legend.position="inside",
        legend.text = element_text(size=16,face = "bold"),
        legend.title = element_text(size=16,face = "bold"),
        axis.title=element_text(size=15))

dev.off()


gutmags %>%
   mutate(posorneg = case_when(manual_ktyp == "K_pos" ~ "K_pos",
                              TRUE ~ "K_neg")) %>%
  dplyr::group_by(posorneg) %>%
  dplyr::summarise(count=n())

```



#### ED7D - MAGs vs genomes (healthy vs disease phylogroups) - only in stool
Same data as 4D (GLM)

```{r}
# ncbi isolates (subset)
ncbi_analysis = read.xlsx(paste0(outputdir,"ncbi_annotated_derep_260112.xlsx")) %>% 
  filter(host_group=="Human" & host_common_name == "Human" & !(isolation_source %in% c("Food","Probiotic","Sewage and wastewater","Meat","Plant"))) %>%
  filter(isolation_source %in% c("Blood", "Brain and CSF") | host_health %in% c("Bloodstream Infection","Meningitis", "neonatal bacteremia","Healthy")) %>%
  mutate(expec = case_when(host_health != "Healthy" ~ 1,
                           TRUE ~ 0)) %>%
  dplyr::ungroup() %>%
  dplyr::select(predicted, host_gender, host_age, mlst, geographic_group, phylogroup,o_type,h_type, expec, group)

# add healthy gut mags
gutmags = readxl::read_xlsx(paste0(outputdir,"gutmags_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           TRUE ~ group),
         expec = 0,
         phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup)) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
  dplyr::ungroup() %>%
   dplyr::select(predicted, host_gender, host_age, mlst, geographic_group, phylogroup,o_type,h_type, expec, group)

# add egli
egli = readxl::read_xlsx(paste0(outputdir,"egli_annotated_metad_ectyper.xlsx")) %>%
  filter(invasiveness == "invasive") %>%
  mutate(expec = 1) %>%
   dplyr::select(predicted, host_gender, host_age, mlst, geographic_group, phylogroup,o_type,h_type, expec, group)


totest = gdata::combine(ncbi_analysis, gutmags, egli) %>%
  filter(phylogroup != "NA") %>%
  mutate(group = case_when(predicted == "K_neg_partial" | group %in% c("No K-type assigned","No KpsC identified") ~ "kps -",
                           TRUE ~ group),
         phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup))


toplot = totest %>%
  dplyr::group_by(phylogroup, expec) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(group,total, phylogroup,expec) %>%
  dplyr::summarise(count = n()) %>%
  mutate(proportion = (count/total)*100)


mycolors = c(
  "G2A" = "#71B7E2",
  "G2B" = "#345C7B",
  "G3A" = "#7653AB",
  "G3B" = "#A48DCC",
  "kps -" = "grey"
)



# Initialize results dataframe
stat_results <- totest %>%
  dplyr::group_by(phylogroup) %>%
  dplyr::summarise(
    p_value = ifelse(n_distinct(group) > 1, # Check if more than one group exists
                     fisher.test(table(group, expec))$p.value, 
                     NA)  # Assign NA if only one group present
  )

# Adjust p-values for multiple comparisons
stat_results$p_adjusted <- p.adjust(stat_results$p_value, method = "bonferroni")

# Convert p-values into significance labels
stat_results$significance <- case_when(
  is.na(stat_results$p_value) ~ "ns",
  stat_results$p_adjusted < 0.001 ~ "***",
  stat_results$p_adjusted < 0.01  ~ "**",
  stat_results$p_adjusted < 0.05  ~ "*",
  TRUE ~ "ns"
)

# Merge significance results into the plotting data
toplot <- left_join(toplot, stat_results, by = "phylogroup")



toplot$phylogroup <- factor(toplot$phylogroup, levels=c("A","B1","C","E1","E2","G","B2-1","B2-2","D1","D2","D3","F"))
stat_results$phylogroup <- factor(stat_results$phylogroup, levels=c("A","B1","C","E1","E2","G","B2-1","B2-2","D1","D2","D3","F"))
toplot$expec <- factor(toplot$expec, levels=c(0,1))

pdf(paste0(plotdir,"ED7D_phylogroup_group_percent_includingKneg_NCBIandcommensals_onlyInvasiveasDisease_plusEgli.pdf"), width = 14, height = 3.5)

ggplot(toplot , aes(x = expec, y = proportion, fill = group, label = count)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = mycolors, guide = guide_legend(ncol = 1), na.value = "grey") +
  geom_text(aes(y = proportion),
            position = position_stack(vjust = 0.5),
            size = 5, 
            color = "black",
            check_overlap = TRUE) + 
  labs(y = "% genomes", x = "", fill = "Capsule Group") +
  
  scale_x_discrete(labels = c("A", "I")) +
  facet_grid(~ phylogroup) +
  theme_classic() +
  theme(axis.text.x = element_text(hjust = 0.5, size = 15),
        axis.title = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        strip.text = element_text(size = 15),
        legend.position = "none") +
  geom_text(aes(x = 1.5, y = 100 + 10, label = significance), 
            size = 5, inherit.aes = FALSE, data = stat_results)

 dev.off()



```


#### ED7E-G - Simpson's diversity, richness and evenness
same data as S5D.
```{r}

# Filter phylogroups with ≥25 genomes in both asymptomatic and invasive
valid_phylogroups <- totest %>%
  count(phylogroup, expec) %>%
  pivot_wider(names_from = expec, values_from = n, values_fill = 0) %>%
  filter(`0` >= 25,`1` >= 25) %>%
  pull(phylogroup)

# Apply filter to main data
totest <- totest %>%
  filter(phylogroup %in% valid_phylogroups)


library(vegan)

# Calculate diversity metrics
diversity_data <- totest %>%
  dplyr::group_by(phylogroup, expec, predicted) %>%
  dplyr::summarise(count = n(), .groups = 'drop') %>%
  pivot_wider(names_from = predicted, values_from = count, values_fill = 0) %>%
  rowwise() %>%
  mutate(
    counts_vec = list(c_across(-c(phylogroup, expec))),
    richness = sum(counts_vec > 0),
    simpson = diversity(counts_vec, index = "simpson"),
    simpson_inverse = diversity(counts_vec, index = "invsimpson"),
    evenness_simpson = ifelse(richness > 1, simpson_inverse / richness, NA)) %>%
  ungroup() %>%
  select(phylogroup, expec, richness, simpson_inverse, evenness_simpson)

# Count unique entries per disease group
n_labels <- totest %>% 
  filter(predicted != "K_neg_partial") %>%
  mutate(phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup)) %>%
  mutate(expec = dplyr::recode(expec, `0` = "A", `1` = "I")) %>%
  dplyr::group_by(expec) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n = ", n))

# Long format for plotting
diversity_long <- diversity_data %>%
  pivot_longer(cols = c(richness, simpson_inverse, evenness_simpson),
               names_to = "metric",
               values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(
    expec = dplyr::recode(expec, `0` = "A", `1` = "I"),
    expec = factor(expec, levels = c("A", "I")),
    metric = factor(metric, levels = c("richness", "simpson_inverse", "evenness_simpson"),
                    labels = c("Richness",
                               "Simpson Diversity Index",
                               "Evenness"))
  )

# Repeat n_labels for each metric panel
metrics <- levels(diversity_long$metric)
n_labels_plot <- tidyr::expand_grid(metric = metrics, n_labels)


num_colors <- n_distinct(diversity_long$phylogroup) 
palette_colors <- colorRampPalette(brewer.pal(12, "Set1"))(num_colors) 

# separate nasty and nice phylogroups
diversity_long = diversity_long %>% 
  mutate(phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup))


pdf(paste0(plotdir,"ED7E_diversitySimpson.pdf"), width =3, height = 2.5)

div_metric = "Simpson Diversity Index"

ggplot(diversity_long %>% filter(metric == div_metric), aes(x = expec, y = value)) +
  geom_boxplot(width = 0.4, outlier.shape = NA, alpha = 0.3,
               position = position_dodge(0.2)) +
  geom_line(aes(group = phylogroup,color = phylogroup),
            position = position_dodge(0.15)) +
  geom_point(aes(group = phylogroup,color = phylogroup), size = 2.5,
             position = position_dodge(0.15)) +
  geom_text_repel(aes(label = phylogroup,color = phylogroup),
                  size = 3.5,
                  position = position_dodge(0.1),
                  show.legend = FALSE,
                  max.overlaps = 20) +
  scale_y_continuous(n.breaks = 5) + 
  scale_color_manual(values=palette_colors) +
  geom_text(data = n_labels_plot,
    aes(x = expec, y = 16, label = label),
    inherit.aes = F, check_overlap = T, 
    #vjust = -3.5,
    size = 4) +
  labs(x = "",y = div_metric) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text=element_text(size=12),
        strip.text = element_blank())

dev.off()



#### Richness
pdf(paste0(plotdir,"ED7F_Richness.pdf"), width = 3, height = 2.5)

div_metric = "Richness"

ggplot(diversity_long %>% filter(metric == div_metric), aes(x = expec, y = value)) +
  geom_boxplot(width = 0.4, outlier.shape = NA, alpha = 0.3,
               position = position_dodge(0.2)) +
  geom_line(aes(group = phylogroup,color = phylogroup),
            position = position_dodge(0.15)) +
  geom_point(aes(group = phylogroup,color = phylogroup), size = 2.5,
             position = position_dodge(0.15)) +
  geom_text_repel(aes(label = phylogroup,color = phylogroup),
                  size = 3.5,
                  position = position_dodge(0.1),
                  show.legend = FALSE,
                  max.overlaps = 10) +
  scale_y_continuous(n.breaks = 5) + 
  scale_color_manual(values=palette_colors) +
  geom_text(
    data = n_labels_plot,
    aes(x = expec, y = 31, label = label),
    inherit.aes = FALSE, check_overlap = T,
    #vjust = -3.5,
    size = 4
  ) +
  labs(
    x = "",
    y = "Richness") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text=element_text(size=12))


dev.off()


pdf(paste0(plotdir,"ED7G_Evenness.pdf"), width = 3, height = 2.5)

div_metric = "Evenness"

ggplot(diversity_long %>% filter(metric == div_metric), aes(x = expec, y = value)) +
  geom_boxplot(width = 0.4, outlier.shape = NA, alpha = 0.3,
               position = position_dodge(0.2)) +
  geom_line(aes(group = phylogroup,color = phylogroup),
            position = position_dodge(0.1)) +
  geom_point(aes(group = phylogroup,color = phylogroup), size = 2.5,
             position = position_dodge(0.1)) +
  geom_text_repel(aes(label = phylogroup,color = phylogroup),
                  size = 3.5,
                  position = position_dodge(0.1),
                  show.legend = FALSE,
                  max.overlaps = 10) +
  scale_y_continuous(breaks = c(0,0.2,0.4,0.6), limits = c(0,0.7)) + 
  scale_color_manual(values=palette_colors) +
  geom_text(
    data = n_labels_plot,
    aes(x = expec, y = 0.65, label = label),
    inherit.aes = FALSE, check_overlap = T,
    size = 4) +
  labs(x = "",y = "Evenness") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text=element_text(size=12))


dev.off()
```



#### ED7H - distribution of Ktypes per isolation source

Same data as 4C.
###### left with K-neg

```{r}

### prep ncbi data
ncbi_analysis = read.xlsx(paste0(outputdir,"ncbi_annotated_derep_260112.xlsx")) %>% 
  filter(host_group == "Human" & host_common_name == "Human" & !(isolation_source %in% c("Food","Probiotic","Sewage and wastewater","Meat","Plant"))) %>%
  mutate(isolation_source = case_when(!(isolation_source %in% c("Stool","Urine and bladder", "Blood","Brain and CSF")) ~ "Other body sites",
                                      TRUE ~ isolation_source),
         host_health = case_when(host_health %in% c("asymptomatic bacteriuria", "Urinary Tract Infection", "recurrent uti") | isolation_source == "Urine and bladder" ~ "UTI",
                                 host_health %in% c("Bloodstream Infection", "neonatal bacteremia") | isolation_source == "Blood" ~ "BSI",
                                 host_health %in% c("Meningitis") | isolation_source == "Brain and CSF" ~ "Meningitis",
                                 host_health %in% c("GI Tract and Diarrhoea", "HUS and TTP", "STEC/Shigella/EHEC") ~ "InPEC",
                                 host_health == "Healthy" & isolation_source == "Stool" ~ "Healthy (stool)",
                                 host_health == "Healthy" & isolation_source != "Stool"~ "Healthy (other)",
                                 TRUE ~ "Other disease")) %>%
  dplyr::ungroup() %>%
  #add genome id column for merging with other datasets - after dereplication we didn't have it anymore
  mutate(genome_id = row_number(),
        group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) %>%
  dplyr::select(genome_id, predicted, host_health,host_age,geographic_group, host_gender, group)


# prep healthy gut MAGs
gutmags = readxl::read_xlsx(paste0(outputdir,"gutmags_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group),
         phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup)) %>%
  dplyr::ungroup() %>%
   dplyr::select(genome_id,predicted, host_health,host_age, host_gender, geographic_group, group) %>%
    mutate(host_health = "MAGs")



# prep egli collection
egli = readxl::read_xlsx(paste0(outputdir,"egli_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group),
         host_health = case_when(invasiveness=="invasive" ~ "BSI",
                                                 TRUE ~ "UTI")) %>%
  dplyr::select(predicted, host_gender, host_health,host_age,geographic_group, group)


common_cols <- intersect(intersect(names(ncbi_analysis), names(egli)), names(gutmags))
totest_invasive <- gdata::combine(ncbi_analysis[common_cols], egli[common_cols], gutmags[common_cols])



leftpanel <- totest_invasive %>%
  group_by(host_health) %>%
  mutate(total = n()) %>%
  ungroup() %>%
  group_by(predicted, total, host_health) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(proportion = (count / total) * 100) %>%
  mutate(predicted = case_when(proportion > 2 ~ predicted,
                               TRUE ~ "Other")) %>%
  group_by(host_health, predicted) %>%
  summarise(count = sum(count), .groups = "drop") %>%
  group_by(host_health) %>%
  mutate(total = sum(count),
         proportion = (count / total) * 100) %>%
  ungroup()

rightpanel <- totest_invasive %>%
  filter(predicted != "K_neg_partial") %>%
  mutate(predicted = gsub("K54_96", "K96", predicted)) %>%
  group_by(host_health) %>%
  mutate(total = n()) %>%
  ungroup() %>%
  group_by(predicted, total, host_health) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(proportion = (count / total) * 100) %>%
  mutate(predicted = case_when(proportion > 4 ~ predicted,
                               TRUE ~ "Other")) %>%
  group_by(host_health, predicted) %>%
  summarise(count = sum(count), .groups = "drop") %>%
  group_by(host_health) %>%
  mutate(total = sum(count),
         proportion = (count / total) * 100) %>%
  ungroup()


#take color mapping from S4B
color_mapping <- readRDS(paste0(outputdir,"Ktype_color_mapping.rds"))
color_mapping <- c(color_mapping[names(color_mapping) %in% unique(c(leftpanel$predicted, rightpanel$predicted))],
                   K_neg_partial = "white")

# Define consistent K-type order
all_predicted <- union(leftpanel$predicted,rightpanel$predicted)
core_ktypes <- setdiff(all_predicted, c("K_neg_partial", "Other"))
predicted_levels <- c(sort(core_ktypes), "Other", "K_neg_partial")

# Apply to both panels
leftpanel$predicted <- factor(leftpanel$predicted, levels = predicted_levels)
rightpanel$predicted <- factor(rightpanel$predicted, levels = predicted_levels)




leftpanel$host_health <- factor(leftpanel$host_health, levels=rev(c("MAGs","Healthy (stool)","Healthy (other)","InPEC","UTI","BSI","Meningitis","Other disease")))


pdf(paste0(plotdir,"ED7H_left_K_type_disease.pdf"), width = 7, height = 3)

totals <- leftpanel %>%
  group_by(host_health) %>%
  summarise(total_count = sum(count), .groups = "drop")

ggplot(leftpanel, aes(x = host_health, y = proportion, fill = predicted, label = count)) +
  geom_bar(stat = "identity", color="grey30",linewidth = 0.15) +
  scale_fill_manual(values = color_mapping) +  
  labs(y = "% genomes", x = "", fill = "K-type") +
  scale_y_continuous(breaks=c(0,25,50,75,100), limits=c(0,110)) +
  geom_text(data = totals, aes(x = host_health, y = 107, label = total_count),
            inherit.aes = FALSE,nudge_y = -3,
            size = 4.5, hjust = 0.2) +
  theme_classic() +
  coord_flip() +
  theme(axis.text.x = element_text(size = 16),
        axis.title = element_text(size = 17),
        axis.text.y = element_text(size = 16),
        strip.text = element_text(size = 17),
        legend.position = "none")
    
dev.off()



```

##### right without K-neg

```{r}

rightpanel$host_health <- factor(rightpanel$host_health, levels=rev(c("MAGs","Healthy (stool)",
                                                                                "Healthy (other)","InPEC",
                                                                                "UTI","BSI","Meningitis",
                                                                                "Other disease")))


pdf(paste0(plotdir,"ED7H_right_K_type_disease_onlyKpos.pdf"), width = 9, height = 3)

# Calculate total counts per bar (disease + phylogroup)
totals <- rightpanel %>%
  group_by(host_health) %>%
  summarise(total_count = sum(count), .groups = "drop")

ggplot(rightpanel, aes(x = host_health, y = proportion, fill = predicted, label = count)) +
  geom_bar(stat = "identity", color="grey30",linewidth = 0.15) +
  scale_fill_manual(values = color_mapping) +  
  scale_y_continuous(breaks=c(0,25,50,75,100), limits=c(0,110)) +
  geom_text(data = totals, aes(x = host_health, y = 107, label = total_count),
            inherit.aes = FALSE,nudge_y = -3,
            size = 4.5, hjust = 0.2) +
  labs(y = expression(paste("% ", italic("kps"), "+ genomes")), x= "", fill="K-type") +
  theme_classic() +
  coord_flip() +
  theme(axis.text.x = element_text(size = 16),
        axis.title = element_text(size = 17),
        axis.text.y = element_text(size = 16),
        strip.text = element_text(size = 17),
        legend.text=element_text(size = 12),
        legend.title=element_text(size = 14),
        legend.key.height = unit(7, "mm"),
        legend.key.width = unit(7, "mm"),
        legend.position = "right") + guides(fill = guide_legend(ncol = 3))



dev.off()



```


## ED Fig. 8

##### prep data - same as fig. 4C
NCBI + Egli + gutmags

```{r}
### prep ncbi data
ncbi_analysis = read.xlsx(paste0(outputdir,"ncbi_annotated_derep_260112.xlsx")) %>% 
  filter(host_group == "Human" & host_common_name == "Human" & !(isolation_source %in% c("Food","Probiotic","Sewage and wastewater","Meat","Plant"))) %>%
  mutate(isolation_source = case_when(!(isolation_source %in% c("Stool","Urine and bladder", "Blood","Brain and CSF")) ~ "Other body sites",
                                      TRUE ~ isolation_source),
         host_health = case_when(host_health %in% c("asymptomatic bacteriuria", "Urinary Tract Infection", "recurrent uti") | isolation_source == "Urine and bladder" ~ "UTI",
                                 host_health %in% c("Bloodstream Infection", "neonatal bacteremia") | isolation_source == "Blood" ~ "BSI",
                                 host_health %in% c("Meningitis") | isolation_source == "Brain and CSF" ~ "Meningitis",
                                 host_health %in% c("GI Tract and Diarrhoea", "HUS and TTP", "STEC/Shigella/EHEC") ~ "InPEC",
                                 host_health == "Healthy" & isolation_source == "Stool" ~ "Healthy (stool)",
                                 host_health == "Healthy" & isolation_source != "Stool"~ "Healthy (other)",
                                 TRUE ~ "Other disease")) %>%
  dplyr::ungroup() %>%
  mutate(genome_id = row_number(),
        group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) 

# prep healthy gut MAGs
gutmags = readxl::read_xlsx(paste0(outputdir,"gutmags_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group),
         phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup)) %>%
  dplyr::ungroup() %>%
    mutate(host_health = "MAGs")


# prep egli collection
egli = readxl::read_xlsx(paste0(outputdir,"egli_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group),
         host_health = case_when(invasiveness=="invasive" ~ "BSI",
                                                 TRUE ~ "UTI"))


common_cols <- intersect(intersect(names(ncbi_analysis), names(egli)), names(gutmags))
totest <- gdata::combine(ncbi_analysis[common_cols], egli[common_cols], gutmags[common_cols])

totest = totest %>%
  filter(mlst !="-")

totest %>% group_by(source) %>%
  summarise(count=n())
```

### ED8A - ST representative vs K-type

```{r}

toplot = totest %>%
  filter(group != "kps -" & phylogroup != "NA" & predicted != "K_neg_partial" & mlst != "-") %>%
  ungroup()

# for text
toplot %>% group_by(source) %>% summarise(n())

toplot =toplot %>%
  dplyr::group_by(predicted) %>%
  dplyr::mutate(no_genomes = n(),
                   no_mlst = n_distinct(mlst)) %>%
  distinct(predicted, group, no_genomes, no_mlst)

mycolors = c(
  "G2A" = "#71B7E2",
  "G2B" = "#345C7B",
  "G3A" = "#7653AB",
  "G3B" = "#A48DCC",
  "kps -" = "grey"
)

pdf(paste0(plotdir,"ED8A_sequencetype_genomes_ktypes_onKtypeside_ncbi_egli_gut.pdf"),width = 4,height = 3)

ggplot(toplot, aes(x=no_genomes,y = no_mlst)) +
  geom_point(aes(fill = group), pch=21,size =2, alpha=0.5) +
  scale_fill_manual(values = mycolors) +
  scale_y_continuous(trans="log2") + 
  geom_text_repel(data = toplot %>% filter(no_genomes > 240),
                  aes(label = predicted, color = group),size = 4, max.overlaps = 10, min.segment.length = 3, fontface="bold") +
  scale_color_manual(values = mycolors) +
  theme_bw() +
  labs(y = "# STs / K-type", x = "#genomes / K-type") +
  guides(alpha="none", fill="none", color="none")

dev.off()

```

### ED8B - capsule diversity within ST (Shannon, Simpson)

```{r}
library(forcats)

div_ST <- totest %>%
  filter(group != "kps -" & phylogroup != "NA" & predicted != "K_neg_partial" & mlst != "-")

div_ST %>% group_by(source) %>%
  summarise(n())

div_ST = div_ST %>%
  count(mlst, predicted) %>%
  group_by(mlst) %>%
  mutate(p = n / sum(n)) %>%
  summarise(
    n_genomes = sum(n),
    shannon = -sum(p * log(p + 1e-12)),
    simpson = 1 - sum(p^2),
    eff_k = exp(shannon),
    .groups = "drop"
  ) %>%
  filter(n_genomes >= 20) %>%
  arrange(desc(shannon)) %>%
  mutate(
    mlst_label = paste0("ST", mlst, " (n=", n_genomes, ")"),
    mlst_label = fct_reorder(mlst_label, shannon)
  )

pdf(paste0(plotdir, "ED8B_ST_capsule_diversity_shannon.pdf"),width = 9,height = 3)

ggplot(div_ST, aes(x = reorder(mlst_label,-shannon), y = shannon)) +
  geom_bar(stat="identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Sequence type (ST)", y = "Shannon diversity (K-type)")

dev.off()



```


### ED8C - ST vs capsule heatmap

```{r}
#remove STs with <20 genomes and kps-neg genomes
st_counts <- totest %>%
  #filter(mlst != "-" & predicted != "K_neg_partial") %>%
  filter(group != "kps -" & phylogroup != "NA" & predicted != "K_neg_partial" & mlst != "-") %>%
  count(mlst, name = "n") %>% arrange(desc(n))

st_keep <- st_counts %>%
  filter(n >= 20) %>%
  pull(mlst)

panelA <- totest %>%
  #filter(mlst != "-" & predicted != "K_neg_partial") %>%
  filter(group != "kps -" & phylogroup != "NA" & predicted != "K_neg_partial" & mlst != "-" & mlst %in% st_keep) 


caps_counts <- panelA %>% count(predicted, name = "n") %>% arrange(desc(n))
caps_keep <- caps_counts %>%
  #here cutoff on number of genoms per capsules - unused
  #filter(n >= 30) %>%
  pull(predicted)

panelA <- panelA %>%
  mutate(capsule_plot = if_else(predicted %in% caps_keep, predicted, "Other"))

panelA %>% group_by(source) %>%
  summarise(n())

# get ST sample sizes for plot
st_n <- panelA %>%
  count(mlst, name = "n") %>%
  mutate(mlst = as.character(mlst))
st_n_vec <- setNames(st_n$n, st_n$mlst)

# row-normalized proportions (ST rows sum to 1)
tab_A <- panelA %>%
  count(mlst, capsule_plot, name = "n") %>%
  group_by(mlst) %>%
  mutate(p = n / sum(n)) %>%
  ungroup()

mat_A <- tab_A %>%
  select(mlst, capsule_plot, p) %>%
  pivot_wider(names_from = capsule_plot, values_from = p, values_fill = 0) %>%
  column_to_rownames("mlst") %>%
  as.matrix()

# remove all-zero rows/cols
mat_A <- mat_A[rowSums(mat_A) > 0, colSums(mat_A) > 0, drop = FALSE]


rownames(mat_A) <- paste0(
  "ST", rownames(mat_A),
  " (n=", st_n_vec[rownames(mat_A)], ")"
)

# Annotate capsules with groups
Kgroups <- readxl::read_xlsx(paste0(inputdir, "subgroups.xlsx"))

caps <- colnames(mat_A)

ha_caps <- HeatmapAnnotation(
  df = data.frame(capsule_plot = caps) %>%
    left_join(Kgroups %>% dplyr::select(-orig_name),
              by = c("capsule_plot" = "new_name")) %>%
    mutate(group = case_when(
      capsule_plot == "K_neg_partial" ~ "K_neg_partial",
      capsule_plot == "Other" ~ "Other",
      TRUE ~ group)) %>%
    distinct(capsule_plot, group) %>%
    dplyr::select(-capsule_plot),
  annotation_height = unit(4, "mm"),
  col = list(group = c(
    "G2A" = "#71B7E2",
    "G2B" = "#345C7B",
    "G3A" = "#7653AB",
    "G3B" = "#A48DCC",
    "K_neg_partial" = "grey",
    "Other" = "grey80"
  )),
  show_legend = TRUE,
  annotation_legend_param = list(
    group = list(direction = "horizontal", nrow = 1)
  )
)

# flip axes
#mat_A_xy <- t(mat_A)
mat_A_xy <- mat_A

# remove any all-zero rows/cols
mat_A_xy <- mat_A_xy[rowSums(mat_A_xy) > 0, colSums(mat_A_xy) > 0, drop = FALSE]

# clustering
hc_cols <- hclust(as.dist(1 - cor(mat_A_xy, method = "spearman")), method = "complete")
dend_cols <- as.dendrogram(hc_cols)

hc_rows <- dendsort(hclust(as.dist(1 - cor(t(mat_A_xy), method = "spearman")), method = "complete"))
dend_rows <- as.dendrogram(hc_rows)


# col_fun <- colorRamp2(
#   c(0, max(mat_A_xy) / 10, max(mat_A_xy)),
#   c("white", "orange", "red")
# )


library(circlize)

# col_fun <- colorRamp2(
#   c(0, max(mat_A_xy)),
#   c("#F7F7F7", "#252525")
# )

col_fun <- colorRamp2(
    c(0, max(mat_A_xy) / 10, max(mat_A_xy)),
    c("white", "orange", "red")
)

pdf(paste0(plotdir, "ED8C_ST_capsule_heatmap.pdf"), width = 8, height = 9)

hm <- Heatmap(
  mat_A_xy,
  name = "proportion",
  col = col_fun,

  cluster_columns = dend_cols,
  cluster_rows    = dend_rows,
  
  top_annotation = ha_caps,

  row_split = 6,
  column_split = 6,

  use_raster = TRUE,
  raster_quality = 5,

  row_title = NULL,
  column_title = NULL,

  rect_gp = gpar(col = "grey85", lwd = 1),
  border = "grey85",

  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),

  heatmap_legend_param = list(
    title = "genomes in K-type\n(proportion)",
    direction = "horizontal"
  )
)

draw(
  hm,
  heatmap_legend_side = "top",
  annotation_legend_side = "top",
  merge_legend = TRUE
)

dev.off()

```

### ED8D - Global association strength (Cramér's V) + permutation null (shuffle capsule within phylogroup)


#### - analysis w/o kps- or partial genomes

```{r}
#remove STs with <20 genomes and kps-neg genomes
st_counts <- totest %>% filter(group != "kps -" & phylogroup != "NA" & predicted != "K_neg_partial" & mlst != "-") %>%
  count(mlst, name = "n") %>% arrange(desc(n))

st_keep <- st_counts %>%
  filter(n >= 20) %>%
  pull(mlst)


panelA <- totest %>%
  #filter(mlst != "-" & predicted != "K_neg_partial") %>%
  filter(group != "kps -" & phylogroup != "NA" & predicted != "K_neg_partial" & mlst != "-" & mlst %in% st_keep) 


caps_counts <- panelA %>% count(predicted, name = "n") %>% arrange(desc(n))
caps_keep <- caps_counts %>%
  #here cutoff on number of genomes per capsules to avoid compromising chisq assumptions with small cells - actually has no impact 
  filter(n >= 5) %>%
  pull(predicted)

panelA <- panelA %>%
  mutate(capsule_plot = if_else(predicted %in% caps_keep, predicted, "Other"))

panelA %>% group_by(source) %>%
  summarise(n())

nrow(panelA)

# Contingency table for ST x capsule_plot
ct_B <- panelA %>%
  mutate(capsule_plot = if_else(predicted %in% caps_keep, predicted, "Other")) %>%
  count(mlst, capsule_plot) %>%
  pivot_wider(names_from = capsule_plot, values_from = n, values_fill = 0) %>%
  column_to_rownames("mlst") %>%
  as.matrix()

# Compute Cramér's V
cramers_v <- function(ct) {
  chi <- suppressWarnings(chisq.test(ct, simulate.p.value = FALSE))
  n <- sum(ct)
  r <- nrow(ct); k <- ncol(ct)
  v <- sqrt(as.numeric(chi$statistic) / (n * (min(r - 1, k - 1))))
  unname(v)
}

V_obs <- cramers_v(ct_B)

# Permute capsules (stratified by phylogroup if available; else global)
permute_capsule <- function(df) {
  if (!all(is.na(df$phylogroup))) {
    df %>%
      group_by(phylogroup) %>%
      mutate(capsule_perm = sample(capsule_plot, size = n(), replace = FALSE)) %>%
      ungroup()
  } else {
    df %>% mutate(capsule_perm = sample(capsule_plot, size = n(), replace = FALSE))
  }
}

df_B <- panelA %>%
  mutate(capsule_plot = if_else(predicted %in% caps_keep, predicted, "Other"),
         genome_id = sprintf("g%06d", dplyr::row_number())) %>%
  select(genome_id, mlst, capsule_plot, phylogroup)

V_perm <- replicate(2000, {
  dfp <- permute_capsule(df_B)
  ct <- dfp %>%
    count(mlst, capsule_perm) %>%
    pivot_wider(names_from = capsule_perm, values_from = n, values_fill = 0) %>%
    column_to_rownames("mlst") %>%
    as.matrix()
  cramers_v(ct)
})

df_V <- tibble(
  V = c(V_obs, V_perm),
  type = c("Observed", rep("Permuted", length(V_perm)))
)

#takes a while to compute so save for fast replotting
write.table(df_V, file=paste0(outputdir,"ED8D_cramersV_permutation_nokps-.tsv"),sep="\t", row.names=FALSE, col.names=TRUE, quote=FALSE)

```


#### - including kps partial

```{r}
#remove STs with <20 genomes and kps-neg genomes
st_counts <- totest %>%
  filter(phylogroup != "NA" & manual_ktyp != "K_neg" & mlst != "-") %>%
  count(mlst, name = "n") %>% arrange(desc(n))

st_keep <- st_counts %>%
  filter(n >= 20) %>%
  pull(mlst)


panelA <- totest %>%
  #filter(mlst != "-" & predicted != "K_neg_partial") %>%
  filter(phylogroup != "NA" & manual_ktyp != "K_neg" & mlst != "-" & mlst %in% st_keep) 


caps_counts <- panelA %>% count(predicted, name = "n") %>% arrange(desc(n))
caps_keep <- caps_counts %>%
  #here cutoff on number of genomes per capsules to avoid compromising chisq assumptions with small cells - actually has no impact 
  filter(n >= 5) %>%
  pull(predicted)

panelA <- panelA %>%
  mutate(capsule_plot = if_else(predicted %in% caps_keep, predicted, "Other"))

panelA %>% group_by(source) %>%
  summarise(n())

nrow(panelA)

# Contingency table for ST x capsule_plot
ct_B <- panelA %>%
  mutate(capsule_plot = if_else(predicted %in% caps_keep, predicted, "Other")) %>%
  count(mlst, capsule_plot) %>%
  pivot_wider(names_from = capsule_plot, values_from = n, values_fill = 0) %>%
  column_to_rownames("mlst") %>%
  as.matrix()

# Compute Cramér's V
cramers_v <- function(ct) {
  chi <- suppressWarnings(chisq.test(ct, simulate.p.value = FALSE))
  n <- sum(ct)
  r <- nrow(ct); k <- ncol(ct)
  v <- sqrt(as.numeric(chi$statistic) / (n * (min(r - 1, k - 1))))
  unname(v)
}

V_obs <- cramers_v(ct_B)

# Permute capsules (stratified by phylogroup if available; else global)
permute_capsule <- function(df) {
  if (!all(is.na(df$phylogroup))) {
    df %>%
      group_by(phylogroup) %>%
      mutate(capsule_perm = sample(capsule_plot, size = n(), replace = FALSE)) %>%
      ungroup()
  } else {
    df %>% mutate(capsule_perm = sample(capsule_plot, size = n(), replace = FALSE))
  }
}

df_B <- panelA %>%
  mutate(capsule_plot = if_else(predicted %in% caps_keep, predicted, "Other"),
         genome_id = sprintf("g%06d", dplyr::row_number())) %>%
  select(genome_id, mlst, capsule_plot, phylogroup)

V_perm <- replicate(2000, {
  dfp <- permute_capsule(df_B)
  ct <- dfp %>%
    count(mlst, capsule_perm) %>%
    pivot_wider(names_from = capsule_perm, values_from = n, values_fill = 0) %>%
    column_to_rownames("mlst") %>%
    as.matrix()
  cramers_v(ct)
})

df_V <- tibble(
  V = c(V_obs, V_perm),
  type = c("Observed", rep("Permuted", length(V_perm)))
)

write.table(df_V, file=paste0(outputdir,"ED8D_cramersV_permutation_withkps-.tsv"),
            sep="\t", row.names=FALSE, col.names=TRUE, quote=FALSE)

```


#### - plot
```{r}

df_nokps <- read.table(paste0(outputdir, "ED8D_cramersV_permutation_nokps-.tsv"),sep = "\t", header = TRUE)

df_withkps <- read.table(paste0(outputdir, "ED8D_cramersV_permutation_withkps-.tsv"),sep = "\t", header = TRUE)

# Extract observed values
V_obs_nokps  <- df_nokps  %>% filter(type == "Observed") %>% dplyr::slice(1) %>% pull(V)
V_obs_withkps <- df_withkps %>% filter(type == "Observed") %>% dplyr::slice(1) %>% pull(V)

# Combine permutation nulls (only Permuted) for overlay
df_perm <- bind_rows(
  df_nokps %>%
    filter(type == "Permuted") %>%
    transmute(V, dataset = "no"),
  df_withkps %>%
    filter(type == "Permuted") %>%
    transmute(V, dataset = "yes")
)

# Observed lines for both
df_obs <- tibble(
  V = c(V_obs_withkps, V_obs_nokps),
  dataset = c("yes", "no"),
  label = paste0("V = ", round(c(V_obs_withkps, V_obs_nokps), 3)),
  y = c(900, 750)  # adjust based on your histogram height
)


cols <- c("no" = "grey70","yes" = "grey20")


pdf(file.path(plotdir, "ED8D_cramersV_overlay_histograms.pdf"),width = 4.5, height = 3)
ggplot(df_perm, aes(x = V, fill = dataset)) +
  geom_histogram(
    bins = 100,
    alpha = 0.5,
    position = "identity",
    color = "grey30"
  ) +
  geom_vline(
    data = df_obs,
    aes(xintercept = V, color = dataset),
    linewidth = 1.1
  ) +
  geom_text(
    data = df_obs,
    aes(x = V, y = y, label = label, color = dataset),
    hjust = 1.1,
    size = 3.2,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = cols, breaks = c("yes", "no")) +
  scale_color_manual(values = cols) +
 # xlim(0.2,0.6) + 
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(x = "Cramer's V",y = "Permutation count",
      # fill = expression(paste("Including ", italic("kps"), "- genomes"))) +
       fill = expression(paste("Including genomes\nwith incomplete K-type loci: "))) +
  guides(color="none")
dev.off()

```


###ED8E - ncbi collection - compare ST similarity according to K-types w/ ST similarity according to genomes (ANI99) or environments
I can do this only for ncbi where ANI99 clusters are available.

##### - prep data

Use NCBI collection bc ANI clusters annotated to assess genetic similarity beyond STs.

```{r}
touse <- ncbi_analysis %>%
  filter(mlst != "-", predicted != "K_neg_partial") %>%
  mutate(mlst = as.character(mlst),
         genome_id = sprintf("g%06d", row_number())) %>%
  # keep only genomes with required ecology fields
  filter(
    !is.na(host_group),
    !is.na(host_common_name),
    !is.na(geographic_group),
    !is.na(isolation_source)
  ) %>%
  mutate(env =interaction(host_group, host_common_name, geographic_group, isolation_source,
                      sep = "_", drop = TRUE)) %>%
  # keep STs with at least 10 genomes
  add_count(mlst, name = "st_n") %>%
  filter(st_n >= 10)

#Filter capsules by number of STs they appear in
caps_keep <- touse %>%
  group_by(predicted) %>%
  summarise(n_st_caps = n_distinct(mlst), .groups = "drop") %>%
  filter(n_st_caps >= 4) %>%
  pull(predicted)

# Filter environments by number of STs they appear in
env_keep <- touse %>%
  group_by(env) %>%
  summarise(n_st_env = n_distinct(mlst), .groups = "drop") %>%
  filter(n_st_env >= 2) %>%
  pull(env)

#Filter ANI99 clusters by number of STs they appear in
ani_keep <- touse %>%
  group_by(ANI99) %>%
  summarise(n_st_ani = n_distinct(mlst), .groups = "drop") %>%
  filter(n_st_ani >= 2) %>%
  pull(ANI99)


touse <- touse %>%
  filter(predicted %in% caps_keep,
         env %in% env_keep,
         ANI99 %in% ani_keep)


nrow(touse)

st_sizes <- touse %>%
  count(mlst, name = "n_genomes") %>%
  arrange(desc(n_genomes))

# calculate jaccard from ST x feature incidence matrix
jaccard_from_incidence <- function(inc_df, st_col = "mlst") {
  stopifnot(!anyDuplicated(inc_df[[st_col]]))
  st_names <- inc_df[[st_col]]
  X <- as.matrix(dplyr::select(inc_df, -all_of(st_col)))
  rownames(X) <- st_names

  X_bin <- X > 0
  A <- X_bin %*% t(X_bin)
  rs <- rowSums(X_bin)
  U <- outer(rs, rs, `+`) - A
  J <- A / U
  diag(J) <- 1
  J
}

# capsule similarity matrix
inc_caps <- touse %>%
  distinct(mlst, predicted) %>%
  mutate(val = 1L) %>%
  tidyr::pivot_wider(names_from = predicted, values_from = val, values_fill = 0L)

J_caps <- jaccard_from_incidence(inc_caps, "mlst")

# environment similarity matrix
inc_env <- touse %>%
  distinct(mlst, env) %>%
  mutate(val = 1L) %>%
  tidyr::pivot_wider(names_from = env, values_from = val, values_fill = 0L)

J_env <- jaccard_from_incidence(inc_env, "mlst")

# ANI99 similarity matrix
inc_gen <- touse %>%
  distinct(mlst, ANI99) %>%
  mutate(val = 1L) %>%
  tidyr::pivot_wider(names_from = ANI99, values_from = val, values_fill = 0L)

J_gen <- jaccard_from_incidence(inc_gen, "mlst")

keep <- st_sizes$mlst
keep <- Reduce(intersect, list(keep, rownames(J_caps), rownames(J_env), rownames(J_gen)))

J_plot <- J_caps[keep, keep, drop = FALSE]
J_env  <- J_env[keep, keep, drop = FALSE]
J_gen  <- J_gen[keep, keep, drop = FALSE]


# ensure no NA/NaN in similarity matrices
stopifnot(all(is.finite(J_plot)), all(is.finite(J_env)), all(is.finite(J_gen)))

# check fraction of zero overlap (should not be ~0 for env anymore)
mean(J_env[upper.tri(J_env)] == 0)
mean(J_gen[upper.tri(J_gen)] == 0)

```


#### - Capsule sharing across STs - similarity matrix

```{r}
edges_bi <- touse %>%
  select(genome_id, mlst, predicted) %>%
  count(mlst, predicted, name = "w")

#Incidence matrix: ST x capsule (0/1)
inc <- edges_bi %>%
  mutate(val = 1L) %>%
  select(mlst, predicted, val) %>%
  pivot_wider(
    names_from  = predicted,
    values_from = val,
    values_fill = 0L
  )

stopifnot(!anyDuplicated(inc$mlst))

st_names <- inc$mlst
X <- as.matrix(select(inc, -mlst))
rownames(X) <- st_names

# Jaccard similarity between ST rows
X_bin <- X > 0
A <- X_bin %*% t(X_bin)                 # intersections
rs <- rowSums(X_bin)
U <- outer(rs, rs, `+`) - A             # unions
J <- A / U
diag(J) <- 1


# Pick top N STs by genomes for heatmap
#TOP_N <- 40
keep <- st_sizes %>%
  arrange(desc(n_genomes)) %>%
  #slice_head(n = TOP_N) %>%
  pull(mlst)

keep <- intersect(keep, rownames(J))
J_plot <- J[keep, keep, drop = FALSE]


```



#### - Ecological similarity

```{r}

# Incidence ST x env
inc_env <- touse %>%
  distinct(mlst, env) %>%
  mutate(val = 1L) %>%
  pivot_wider(names_from = env, values_from = val, values_fill = 0L)

stopifnot(!anyDuplicated(inc_env$mlst))

st_names <- inc_env$mlst
X <- as.matrix(select(inc_env, -mlst))
rownames(X) <- st_names

# Jaccard similarity
X_bin <- X > 0
A <- X_bin %*% t(X_bin)
rs <- rowSums(X_bin)
U <- outer(rs, rs, `+`) - A
J_env <- A / U
diag(J_env) <- 1

# keep same ST ordering as your capsule/gen matrices
keep <- st_sizes %>% arrange(desc(n_genomes)) %>% pull(mlst)
keep <- intersect(keep, rownames(J_env))
J_env <- J_env[keep, keep, drop = FALSE]

```

####- Genetic similarity (ANI99)

```{r}
edges_bi <- touse %>%
  select(genome_id, ANI99, mlst) %>%
  count(mlst, ANI99, name = "w")

#Incidence matrix: ST x capsule (0/1)

inc <- edges_bi %>%
  mutate(val = 1L) %>%
  select(mlst, ANI99, val) %>%
  pivot_wider(
    names_from  = ANI99,
    values_from = val,
    values_fill = 0L
  )

stopifnot(!anyDuplicated(inc$mlst))

st_names <- inc$mlst
X <- as.matrix(select(inc, -mlst))
rownames(X) <- st_names

# Jaccard similarity between ST rows

X_bin <- X > 0
A <- X_bin %*% t(X_bin)                 # intersections
rs <- rowSums(X_bin)
U <- outer(rs, rs, `+`) - A             # unions
J <- A / U
diag(J) <- 1


# Pick top N STs by genomes for heatmap
#TOP_N <- 40
keep <- st_sizes %>%
  arrange(desc(n_genomes)) %>%
  #slice_head(n = TOP_N) %>%
  pull(mlst)

keep <- intersect(keep, rownames(J))
J_gen <- J[keep, keep, drop = FALSE]

```


#### - Compare the Jaccard matrices (capsule vs env or vs gen)
```{r}

# compare off-diagonal entries only
get_offdiag <- function(M) M[upper.tri(M)]

# capsule vs environment
cor(get_offdiag(J_plot), get_offdiag(J_env), method = "spearman")
mantel(as.dist(1 - J_plot), as.dist(1 - J_env), method = "spearman", permutations = 9999)

# capsule vs ANI99
cor(get_offdiag(J_plot), get_offdiag(J_gen), method = "spearman")
mantel(as.dist(1 - J_plot), as.dist(1 - J_gen), method = "spearman", permutations = 9999)


```

#### - Figure

```{r}

# Capsule vs ecology
mantel_env <- mantel(
  as.dist(1 - J_plot),
  as.dist(1 - J_env),
  method = "spearman",
  permutations = 9999
)

# Capsule vs genome
mantel_gen <- mantel(
  as.dist(1 - J_plot),
  as.dist(1 - J_gen),
  method = "spearman",
  permutations = 9999
)


# Combine permutation nulls
df_mantel <- bind_rows(
  tibble(rho = mantel_env$perm, comparison = "Capsule vs ecology"),
  tibble(rho = mantel_gen$perm, comparison = "Capsule vs 99% ANI")
)

# Observed statistics
df_obs <- tibble(
  rho = c(mantel_env$statistic, mantel_gen$statistic),
  comparison = c("Capsule vs ecology", "Capsule vs 99% ANI")
)


#write these down to avoid recomputing
write.table(df_mantel, file=paste0(outputdir,"ED6E_mantel_permutation.tsv"),sep="\t", row.names=FALSE, col.names=TRUE, quote=FALSE)

write.table(df_obs, file=paste0(outputdir,"ED6E_mantel_observed.tsv"),sep="\t", row.names=FALSE, col.names=TRUE, quote=FALSE)

#read these back
df_mantel <- read.table(paste0(outputdir,"ED6E_mantel_permutation.tsv"),sep = "\t", header = TRUE)
df_obs <- read.table(paste0(outputdir,"ED6E_mantel_observed.tsv"),sep = "\t", header = TRUE)


pdf(paste0(plotdir, "ED8E_mantel_overlay_histograms.pdf"), width = 4.5, height = 3)

ggplot(df_mantel, aes(x = rho, fill = comparison)) +
  geom_histogram(aes(fill=comparison),bins = 100,alpha = 0.5,position = "identity") +
  geom_vline(data = df_obs,
             aes(xintercept = rho, color = comparison),
             linewidth = 1.1) +
  geom_text(data = df_obs, aes(x = rho+0.15,y = 450,label = paste0("r = ", round(rho, 3)),
                               color = comparison),
            vjust = -0.4,hjust = 1.1,size = 3.2,show.legend = FALSE) +
  # scale_fill_manual(values = c("Capsule vs ecology" = "#1B9E77",
  #                              "Capsule vs 99% ANI" = "#D95F02")) +
  # scale_color_manual(values = c("Capsule vs ecology" = "#1B9E77",
  #                               "Capsule vs 99% ANI" = "#D95F02")) +
  scale_fill_manual(values = c("Capsule vs ecology" = "orange",
                               "Capsule vs 99% ANI" = "grey10")) +
  scale_color_manual(values = c("Capsule vs ecology" = "orange",
                                "Capsule vs 99% ANI" = "grey10")) +
  theme_bw() +
  labs(x = "Mantel Spearman's r",
       y = "Permutation count",
       fill="ST similarity comparison") +
  guides(color="none") +
  theme(legend.position = "bottom", legend.title.position = "top", legend.title = element_text(hjust=0.5))

dev.off()

```



## ED Fig. 9

##### load annotation O-types
```{r}
processed_sugars = fread(paste0(inputdir,"Final_Processed_O-Antigen_Data.csv")) %>%
  clean_names() %>%
  mutate(o_type = sub(" .*", "", o_antigen)) %>%  # remove after space
  distinct(o_type, .keep_all = TRUE) %>%
  mutate(sugar = case_when(o_type == "O25" ~ "aDGlc|aLFucNAc|bDGlcNAc|aLRha",
                           TRUE ~ sugar))

```

##### prep data - same as fig. 4C
NCBI + Egli + gutmags

```{r}
### prep ncbi data
ncbi_analysis = read.xlsx(paste0(outputdir,"ncbi_annotated_derep_260112.xlsx")) %>% 
  filter(host_group == "Human" & host_common_name == "Human" & !(isolation_source %in% c("Food","Probiotic","Sewage and wastewater","Meat","Plant"))) %>%
  mutate(isolation_source = case_when(!(isolation_source %in% c("Stool","Urine and bladder", "Blood","Brain and CSF")) ~ "Other body sites",
                                      TRUE ~ isolation_source),
         host_health = case_when(host_health %in% c("asymptomatic bacteriuria", "Urinary Tract Infection", "recurrent uti") | isolation_source == "Urine and bladder" ~ "UTI",
                                 host_health %in% c("Bloodstream Infection", "neonatal bacteremia") | isolation_source == "Blood" ~ "BSI",
                                 host_health %in% c("Meningitis") | isolation_source == "Brain and CSF" ~ "Meningitis",
                                 host_health %in% c("GI Tract and Diarrhoea", "HUS and TTP", "STEC/Shigella/EHEC") ~ "InPEC",
                                 host_health == "Healthy" & isolation_source == "Stool" ~ "Healthy (stool)",
                                 host_health == "Healthy" & isolation_source != "Stool"~ "Healthy (other)",
                                 TRUE ~ "Other disease")) %>%
  dplyr::ungroup() %>%
  #add genome id column for merging with other datasets - after dereplication we didn't have it anymore
  mutate(genome_id = row_number(),
        group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group)) #%>%
 # dplyr::select(genome_id, predicted, host_health,host_age,geographic_group, host_gender, group)

# prep healthy gut MAGs
gutmags = readxl::read_xlsx(paste0(outputdir,"gutmags_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted %in% c("No KpsC identified", "No K-type assigned") ~ predicted,
                           predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group),
         phylogroup = case_when(phylogroup == "C_original" ~"C",
                                TRUE ~ phylogroup)) %>%
  dplyr::ungroup() %>%
  # dplyr::select(genome_id,predicted, host_health,host_age, host_gender, geographic_group, group) %>%
    mutate(host_health = "MAGs")


# prep egli collection
egli = readxl::read_xlsx(paste0(outputdir,"egli_annotated_metad_ectyper.xlsx")) %>%
  mutate(group = case_when(predicted == "K_neg_partial" ~ "kps -",
                           TRUE ~ group),
         host_health = case_when(invasiveness=="invasive" ~ "BSI",
                                                 TRUE ~ "UTI")) #%>%
  #dplyr::select(predicted, host_gender, host_health,host_age,geographic_group, group)


common_cols <- intersect(intersect(names(ncbi_analysis), names(egli)), names(gutmags))
totest <- gdata::combine(ncbi_analysis[common_cols], egli[common_cols], gutmags[common_cols])

totest = totest %>%
  filter(mlst !="-") %>%
  mutate(mlst = gsub("\\.0","",mlst))

#-------- harmonise O types
# Function to sort components of serotype strings
harmonize_serotype <- function(x) {
  if (grepl("/", x)) {
    parts <- unlist(strsplit(x, "/"))
    parts <- sort(parts)
    return(paste(parts, collapse = "/"))
  } else {
    return(x)
  }
}

totest$o_type <- sapply(totest$o_type, harmonize_serotype)

totest %>% group_by(source) %>%
  summarise(count=n())

```


### ED9A - K-type/O-type proportions


```{r}

# filter and assign rare h_types as "Other"
totest_filtered <- totest %>%
  filter(manual_ktyp == "K_pos") %>%
  group_by(o_type) %>%
  mutate(no_genomes = n()) %>%
  ungroup() %>%
  filter(no_genomes > 25) %>%
  group_by(o_type, h_type) %>%
  mutate(frac = n() / dplyr::first(no_genomes)) %>%
  ungroup() %>%
  mutate(h_type = ifelse(frac < 0.3, "Other", h_type))


# assign each (o_type, K-type) to one h_type only
k_assignments <- totest_filtered %>%
  distinct(o_type, predicted, h_type) %>%
  dplyr::group_by(o_type, predicted) %>%
  dplyr::slice(1) %>%  # Assign each K-type to just one H-type per O-type
  ungroup()

# count unique K-types per (o_type, h_type)
toplot <- k_assignments %>%
  group_by(o_type, h_type) %>%
  summarise(unique_ktypes = n(), .groups = "drop")


# get genome counts per o_type for bar labels
bar_labels <- totest_filtered %>%
  dplyr::group_by(o_type) %>%
  dplyr::summarise(no_genomes = n(), .groups = "drop") %>%
  left_join(
    toplot %>%
      group_by(o_type) %>%
      summarise(total_height = sum(unique_ktypes), .groups = "drop"),
    by = "o_type"
  )


h_types <- unique(toplot$h_type)
non_other_h <- setdiff(h_types, "Other")
palette_colors <- hcl.colors(length(non_other_h), palette = "Dynamic")
mycolors <- setNames(c(palette_colors, "grey"),
                     c(non_other_h, "Other"))

toplot <- toplot %>%
  left_join(bar_labels %>% select(o_type, total_height), by = "o_type")


# upper panel
pdf(paste0(plotdir,"ED9A_upperpanel_OKtype_barplot_normgenomes.pdf"),width=10,height=5)
ggplot(toplot, aes(x = reorder(o_type, -total_height),
                   y = unique_ktypes, fill = h_type)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(data = bar_labels,
            aes(x = o_type, y = total_height + 4, label = no_genomes),
            inherit.aes = FALSE,
            angle = 90, vjust = 0.5, size = 3) +
  scale_fill_manual(values = mycolors,
                    guide = guide_legend(ncol = 2)) +
  labs(x = "O-type",
       y = "Unique K-types per O-type",
       fill = "H-type") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        axis.title = element_text(size = 10))

dev.off()


# lower panel (kps+ / - )
toplot = toplot %>% 
  mutate(o_type = reorder(o_type, -total_height))
desired_order <- levels(toplot$o_type)


toplot = totest %>%
  mutate(manual_ktyp = case_when(manual_ktyp == "K_neg" ~ "-",
                             TRUE ~ "+")) %>%
  filter(o_type %in% toplot$o_type) %>%
  dplyr::group_by(o_type) %>%
  dplyr::mutate(total=n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(total, o_type, manual_ktyp) %>%
  dplyr::summarise(count = n()) %>%
  ungroup() %>%
  mutate(proportion = (count/total)*100)



toplot$o_type <- factor(toplot$o_type, levels = desired_order)


pdf(paste0(plotdir,"ED9A_lowerpanel_O_type_kps+.pdf"),width = 10,height=5)
ggplot(toplot, aes(x = o_type, y = proportion, fill = manual_ktyp, label = count)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(y = proportion),
            position = position_stack(vjust = 0.5),angle = 90,
            size = 3, 
            color = "black",
            check_overlap = TRUE) + 
  scale_fill_manual(values = c("white", "grey"), labels = c("-", "+")) +
 # coord_flip() +
  theme_classic() +
  theme(
    axis.text.x = element_text(hjust = 1,angle =60),
    axis.title = element_text(size = 10),
    legend.position = "right") +
  labs(y = "% genomes", x = "", fill = "Capsule")


dev.off()



```

###ED9B - O-type-K-type heatmap


```{r}
test = totest %>%
  filter(manual_ktyp == "K_pos" &  o_type != "-") 

nrow(test)

test = test %>%
  dplyr::group_by(predicted) %>%
  dplyr::mutate(total = n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(predicted, o_type, total) %>%
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::mutate(proportion = count / total) %>%
  distinct(predicted, o_type, proportion) %>%
  pivot_wider(names_from = predicted, values_from = proportion)

test[is.na(test)] <- 0

mat_data = test %>%
  as.data.frame() %>%
  column_to_rownames("o_type") %>%
  as.matrix()

# Remove rows/columns with all 0
mat_data <- mat_data[rowSums(mat_data != 0) > 0, colSums(mat_data != 0) > 0]

# dendrograms
hc <- hclust(as.dist(1 - cor(mat_data, method = "spearman")), method = "complete")
hr <- dendsort(hclust(as.dist(1 - cor(t(mat_data), method = "spearman")), method = "complete"))

dend1 <- as.dendrogram(hc)
type <- colnames(mat_data)

# k-type group annotation
ha = HeatmapAnnotation(
  df = data.frame(predicted = type) %>%
    left_join(Kgroups %>% dplyr::select(-orig_name), by = c("predicted" = "new_name")) %>%
    mutate(group = case_when(predicted == "K_neg_partial" ~ "K_neg_partial", TRUE ~ group)) %>%
    distinct(predicted, group) %>%
    dplyr::select(-predicted),
  annotation_height = unit(4, "mm"),
  col = list(group = c("G2A" = "#71B7E2", "G2B" = "#345C7B", "G3A" = "#7653AB", "G3B" = "#A48DCC", "K_neg_partial" = "grey")),
   show_legend = TRUE
)

# sugar annotation

# get brooader sugar groups (Tim's version abbreviations)
group_definitions <- list(
  Rha = c("aLRha", "bLRha"),
  GlcNAc = c("bDGlcNAc", "aDGlcNAc"),
  Gal = c("aDGal", "bDGal"),
  GalNAc = c("aDGalNAc", "bDGalNAc"),
  Glc = c("aDGlc", "bDGlc"),
  Man = c("aDMan", "bDMan"),
  Fuc = c("aLFuc", "bLFuc"),
  HexA = c("GlcA", "GalA"),
  Rare_sugars = c("QuiNAc", "Tyv", "Abe")
)

# assign groups based on sugars
assign_sugar_groups <- function(sugar_string) {
  sugars <- unlist(strsplit(sugar_string, "\\|"))
  matched_groups <- sapply(names(group_definitions), function(group) {
    if (any(sugars %in% group_definitions[[group]])) return(group) else return(NULL)
  })
  paste(na.omit(unlist(matched_groups)), collapse = "; ")
}

processed_sugars$sugar_group <- sapply(processed_sugars$sugar, assign_sugar_groups)

sugar_data <- processed_sugars %>%
   dplyr::select(o_type, sugar_group) %>%
   rename(sugar = sugar_group)


# prep sugar abbreviated annotation 
sugar_ann <- data.frame(o_type = rownames(mat_data)) %>%
  left_join(sugar_data, by = "o_type") %>%
  mutate(sugar = ifelse(is.na(sugar), "Unknown", sugar)) %>%
# some are not present bc ECtyper gives comprehensive annotation not present in this document --> integrate
  mutate(sugar=case_when(grepl("O106", o_type) ~ "GlcN; Man",
                          o_type == "O107/O117" ~ "Rha; GlcN; Gal",
                        o_type == "O13/O135" ~ 	"Rha; GlcN",
                        o_type == "O2/O50" ~ 	"Rha; GlcN",
                        o_type == "O142/O157" ~ 	"Rha; Gal; Glc; Fuc",
                         o_type == "O62" ~ "D-Galf",
                        o_type == "O89" ~ "Unknown",
                        TRUE ~ sugar))


unique_sugars <- unique(sugar_ann$sugar)
sugar_colors <- setNames(rainbow(length(unique_sugars)), unique_sugars)

library(viridis)
sugar_colors <- setNames(viridis(length(unique_sugars), option = "turbo"), unique_sugars)

sugar_color_anno <- rowAnnotation(
  sugar_color = sugar_ann$sugar,
  col = list(sugar_color = sugar_colors),
  annotation_width = unit(4, "mm"),
  show_legend = TRUE
)

sugar_label_anno <- rowAnnotation(
  sugar_label = anno_text(sugar_ann$sugar, just = "left", gp = gpar(fontsize = 8)),
  annotation_width = unit(30, "mm")
)

rownames(sugar_ann) <- sugar_ann$o_type
sugar_ann <- sugar_ann[rownames(mat_data), ]

row_ha <- rowAnnotation(
  sugar_color = sugar_ann$sugar,
  sugar_label = anno_text(sugar_ann$sugar, just = "left", gp = gpar(fontsize = 11)),
  col = list(sugar_color = sugar_colors),
  annotation_width = unit.c(unit(4, "mm"), unit(50, "mm")),
  show_annotation_name = FALSE,
  show_legend = c(sugar_color = FALSE) 
)


col_fun <- colorRamp2(
  c(0, max(mat_data) / 10, max(mat_data)),
  c("white", "orange", "red")
)

# plot
pdf(paste0(plotdir, "ED9B_clustered_OK.pdf"), width = 19, height = 16)


hm <- Heatmap(mat_data,
        cluster_columns = dend1,
        col = col_fun,
        top_annotation = ha,
        left_annotation = row_ha,
        row_split = 10, column_split = 6,
        use_raster = TRUE, raster_quality = 5,
        row_title = NULL,
        column_title = NULL,
        border = "grey80",
        rect_gp = gpar(col = "grey90", lwd = 3),
        heatmap_legend_param = list(title = "proportion genomes\nin K-type"))


draw(hm,
     heatmap_legend_side = "right",
     annotation_legend_side = "right",
     merge_legend = TRUE)


dev.off()


```


##### for text

```{r}
harmonize_serotype <- function(x) {
  if (grepl("/", x)) {
    parts <- unlist(strsplit(x, "/"))
    parts <- sort(parts)
    return(paste(parts, collapse = "/"))
  } else {
    return(x)
  }
}

totest$o_type <- sapply(totest$o_type, harmonize_serotype)

test = totest %>%
  filter(predicted != "K_neg_partial" & o_type != "-") %>%
  dplyr::group_by(predicted) %>%
  dplyr::mutate(total = n()) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(predicted, o_type, total) %>%
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::mutate(proportion = count / total) %>%
  distinct(predicted, o_type, proportion) 

# Assign sugar groups
assign_sugar_groups <- function(sugar_string) {
    sugars <- unlist(strsplit(sugar_string, "\\|"))
    matched_groups <- sapply(names(group_definitions), function(group) {
        if (any(sugars %in% group_definitions[[group]])) return(group) else return(NULL)
    })
    paste(na.omit(unlist(matched_groups)), collapse = "; ")
}

processed_sugars$sugar_group <- sapply(processed_sugars$sugar, assign_sugar_groups)




# Align sugar annotations
test <- test %>%
    left_join(processed_sugars, by = "o_type") %>%
  mutate(sugar_group = case_when(
        grepl("O106", o_type) ~ "Glucosamine; Mannose",
        o_type == "O107/O117" ~ "Rhamnose; Glucosamine; Galactose",
        o_type == "O13/O135" ~ "Rhamnose; Glucosamine",
        o_type == "O2/O50" ~ "Rhamnose; Glucosamine",
        o_type == "O142/O157" ~ "Rhamnose; Galactose; Glucose; Fucose",
        o_type == "O62" ~ "D-galactofuranose",
        o_type == "O89" ~ "Unknown",
        TRUE ~ sugar_group
    ))


    

```



